---
title: "R Fundamentals Part 4 Project"
author: "D-Lab"
date: "9/27/2018"
output: 
  html_document: 
    toc: yes
    toc_float: yes
---

# Example workflow

You are given a dataset you have never seen before. You want to use RStudio to import the dataset and learn about it a little before you proceed with preprocessing tasks, exploratory data analysis, and statistical testing. 

##### 1. Load the `mtcars` dataset.  
```{r}
## YOUR CODE HERE
```

##### 2. What are some useful functions to learn more about the data? 
```{r}
## YOUR CODE HERE
```

##### 3. Might you surmise a relationship about something like engine size and miles per gallon?  
```{r}
## 
```

##### 4. Create boxplots for mpg as parsed by cyl. 
```{r}
## YOUR CODE HERE
```

##### 5. What does `cor.test` reveal about mpg and cyl?  
```{r}
## YOUR CODE HERE
```

##### 6. Create a scatterplot of the "mpg" and "hp" variables using `ggplot()`. What do you notice?  
```{r}
## YOUR CODE HERE
```

##### 7. Can cyl be used to predict mpg in a linear regression model? Why must you learn about the assumptions of a statistical test before applying it (or any other statistical test)? 
```{r}
## YOUR CODE HERE
```
  
# Gapminder dataset
### 1. 
Now that you are warmed up, load the "gapminder-FiveYearData.csv" file for yet another dataset you have never seen before. Save it in a variable named `gap`:
```{r}
## YOUR CODE HERE
```

### 2. 
How would you learn more about this dataset? 
```{r}
## YOUR CODE HERE
```

### 3. 
Accomplish the following tasks: 

### A) Write three sentences describing this dataset and write one sentence that proposes a research question to be investigated. 
```{r}
## 
## 
## 

## 
```

### B) Turn your research question into a testable hypothesis:
```{r}
## 
```

### C) Compute descriptive statistics for each variable in the dataset.  
```{r}
## YOUR CODE HERE
```

### D) Can you visualize the data from every single column in this dataset? Create a histogram, boxplot, or scatterplot for the ones you can, and explain in writing the ones you cannot.  
```{r}
## YOUR CODE HERE

## 
```

### E) Do these plots inform you about anything specific that fits your hypothesis and that you can justifiably test? 
```{r}
## 
##
##
```

### F) Interpret your plots and your statistical tests - what is going on? 
```{r}
##
##
##
```

