---
title: 'R Fundamentals Part 2: Subsetting and reshaping'
author: "Evan Muzzall, Shinhye Choi, Rochelle Terman, Dillon Niederhut"
date: "February 20, 2016"
output:
  html_document:
    toc: yes
    toc_float: yes
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 2 Learning objectives
1. Day 1 review
2. Loading data from files
3. Subsetting in base R
4. Missing data (NA)
5. Merging data
6. Subsetting with the dplyr R package
7. Tidying/reshaping data with the tidyr R package

# 1.  Day 1 review
1. Set your working directory
2. The assignment operator `<-`
3. Atomic data types: numeric, character, logical
4. Data structures: vector, list, matrix, data frame
5. Save your work: `write.csv()` and `sink()`

# 2.  Loading data from files
Set your working directory to the R-Fundamentals data folder:
```{r, eval=FALSE}
getwd()
setwd("/Users/evanmuzzall/Desktop/R-Fundamentals")
```
Now, load the animals dataset that you created in Part 1. It is conveniently located in the "data" folder. Note that we can also specify the file path within `read.csv`:
```{r}
?read.csv
```
```{r}
animals <- read.csv("./data/animals.csv", header=TRUE, stringsAsFactors=FALSE)

# . means "current folder"
# .. means "folder above"

# or
animals <- read.csv("/Users/evanmuzzall/Desktop/R-Fundamentals/data/animals.csv", header=TRUE, stringsAsFactors = FALSE)
```
> Notice that `stringsAsFactors=FALSE`. If set to `TRUE`, R will try to guess which character data vectors should automatically be converted to factors. This is problematic because 1) R is not always good at guessing and 2) R defaults to alphabetical factor level sorting. This might not matter for your data, but we recommend to set `stringsAsFactors=FALSE` and manually convert your desired character vectors to factors. Refer back to the end of Part 1 for these instructions. `header=TRUE` will include the header row; `header=FALSE` will turn your header row into the first row of actual data.  

Also load the mammals "sleep" dataset from the [VIM package](https://cran.r-project.org/web/packages/VIM/VIM.pdf). This dataset looks at sleep patterns in various mammal species and contains missing (`NA`) values. See the link below for the original research article:
```{r}
library(VIM)
data(sleep) 
```
or, to read it in from the data folder ...
```{r, eval=FALSE}
sleep <- read.csv("./data/sleep_VIM.csv", header=TRUE, stringsAsFactors=FALSE)
```
View the original [Allison and Chichetti 1976 article here](http://people.stat.sfu.ca/~cschwarz/Stat-650/Notes/MyPrograms/MultReg/AnimalSleep/AllisonScience1976.pdf)

> NOTE: you DO NOT want to load the Student's Sleep Dataset from the "datasets" library. You should have 62 rows and 10 columns.  

```{r, eval=FALSE}
?sleep # Click the "Mammal sleep data" link. You DO NOT want "Student's Sleep Data".
```

Remember, we can save this dataframe to a .CSV file with `write.csv()`. It will save to our working directory:
```{r, eval=FALSE}
?write.csv
```
```{r}
write.csv(sleep, "sleep_VIM.csv", row.names=FALSE)
```
`row.names=` will specify if you want to add a column to the .CSV file with the row names. Let's say no by giving this parameter a `FALSE` argument.  

Like above, we can then use `read.csv` to load it into R. However, note that here we are including the file path in our `read.csv` function. This way, you can load data from other folders without changing your working directory! 
```{r, eval=FALSE}
sleep <- read.csv("/Users/evanmuzzall/Desktop/R-Fundamentals/data/sleep_VIM.csv", header=TRUE, stringsAsFactors=FALSE)

# or 

getwd()
setwd("/Users/evanmuzzall/Desktop/R-Fundamentals/data")
sleep <- read.csv("sleep_VIM.csv", header=TRUE, stringsAsFactors=FALSE)
```

> NOTE: the main point to remember is that you should first set your working directory to the folder that contains the data you wish to import. Whether you specify this using setwd() or within the read.csv() function is up to you. However, if you wish to import data from multiple locations you might find it easier to specify the path within read.csv so that you do not have to change your working directory many times.  

When dealing with Microsoft Excel files (.XLSX), you might find it more convenient to save them first as .CSV files in Excel and then import them using `read.csv()`.  

Other functions also work to import data from files, such as `load()`. You might also have success with the "xlsx" R package and its `read.xlsx()` command for directly importing Excel files.  

Also, the "foreign" R package has commands for loading data from SAS, SPSS, Stata, etc.

# 2.  Loading data from files - inspecting the data frame
Remember from Part 1 that we can learn a lot about data in R. For dataframes, the following commands are common:
```{r, eval=FALSE}
str(sleep)    # returns the structure of the dataframe
dim(sleep)    # dataframe dimensions; remember, we want 62 rows and 10 columns! 

rownames(sleep)   #row names (they have not been named and default to character type)
# for example:
class(rownames(sleep))

nrow(sleep)   # number of rows
ncol(sleep)   # number of columns
unique(sleep)   # show rows with unique data

head(sleep, n=10) # show the first "n" rows
tail(sleep, n=4) # show the last "n" rows
```
`names()` and `colnames()` both return column names of the data frame:
```{r}
names(sleep)  
colnames(sleep)
```
We can also check which indices are true. Let's convert the "Span" column to a logical vector where missing data is coded as `NA` while present data is marked `TRUE`.
```{r, eval=FALSE}
?as.logical
```
```{r}
new_Span <- as.logical(sleep$Span)
new_Span
```
Now we can see which data are missing (`NA`) and which cells have data present (`TRUE`).

`which()` will return the rows numbers that have data present:
```{r, eval=FALSE}
?which
```
```{r}
which(new_Span)
```

# 3.  Subsetting in base R
Efficiently subsetting your data will save you time and stress. Fortunately, there are several different ways to subset data in base R.

##  3.  Subsetting in base R - `$`
Remember from Part 1 that the dollar sign operator `$` will extract only a single vector/column within the data frame:
```{r, eval=FALSE}
?"$" # Remember that we must wrap symbols in quotation marks to view their help pages
```
```{r}
sleep$Dream #Returns only the "Dream" vector from the `sleep` data frame.
```
However, you might find subsetting using **bracket notation** `[ , ]` along with variable names, positive and negative integers, and/or logical values is easier because you can subset multiple elements at once. 

##  3.  Subsetting in base R - `[,c(Variable Names)]`
You can subset your data by specifying variable names within bracket notation and using the `c()` command to create a column name character vector of names you want to keep.

We can create a new dataframe object `sleep_varnames` that includes only "BodyWgt" and "BrainWgt" variable names from the `sleep` data frame:
```{r, eval=FALSE}
?"["
```
```{r}
library(VIM)
data(sleep)
sleep_varnames <- sleep[,c("BodyWgt", "BrainWgt")]
str(sleep_varnames)
head(sleep_varnames)
```

> Notice that the comma is still included within the bracket notation before the vector of column names. This indicates that we want ALL of the rows corresponding to these two columns. This is the same when we only want to subset rows and include ALL columns (see below). 

##  3.  Subsetting in base R - two-dimensional subsetting `[c(x:y), c(x:y)]`
When you subset your data in two dimensions, you subset both the rows and columns.

Remember that in bracket notation `[ , ]` everything **before** the comma refers to rows, and everything **after** the comma refers to columns!

##  3.  Subsetting in base R - logical tests
We can also use logical tests to subset our data. For example, what if we want to include only the rows that have a value of 1 for "Exp"? We can use the relational operator `==`:
```{r, eval=FALSE}
?"=="
```
```{r}
sleep_logical <- sleep[sleep$Exp == 1,]
sleep_logical
```
Only rows with values of 1 for "Exp" are returned! 

What if we want to subset the data so it includes only the rows that have a value of 1 **and** 3 for Exp and 2 for "Danger"? We would use the "and" `&` logical operator:
```{r, eval=FALSE}
?"&"
```
```{r}
sleep_logical2 <- sleep[sleep$Exp == c(1,3) & sleep$Danger == 2,]
sleep_logical2
```
We can subset a dataframe that includes only rows with "Exp" = 2 **and** 3 and only "BodyWgt", "BrainWgt", and "Exp" columns. 
```{r, eval=FALSE}
str(sleep)
```
```{r}
sleep_2d <- sleep[sleep$Exp == c(2,3), c("BodyWgt", "BrainWgt", "Exp")]
sleep_2d
```

> NOTE: see how we have the column names listed after the comma? When we only want to call rows, we still must include the comma to tell R that we want ALL of the columns! For example, if we want to specify only rows that have values of 2 and 3 for "Exp" and want all of the columns, we still must include the column inside our bracket notation after the specified rows:

```{r}
sleep_rows_only <- sleep[sleep$Exp == c(2,3),]
sleep_rows_only
```

##  3.  Subsetting in base R - subsetting with positive integers `[,c(x:y)]`
Subsetting by **positive** integers works as well. This will **include** only the column numbers specified, rather than typing out their names.

Let's create an object `sleep_posint` that includes only "NonD", "Sleep", and "Exp" columns. 

First use `str()` to see which integer values these columns represent. Because we only want the 3rd, 5th, and 9th columns, we type:
```{r, eval=FALSE}
str(sleep)
sleep_posint <- sleep[,c(3,5,9)]
str(sleep_posint)
head(sleep_posint)
```

##  3.  Subsetting in base R - subsetting with negative integers `[,-c(x:y)]`
Subsetting by **negative** integers will **exclude** the specified columns. Notice the `-` symbol before `c()` inside our bracket notation.

We can create an object called `sleep_negint` that includes everything **except** columns 1 and 2 ("BodyWgt" and "BrainWgt"):
```{r}
str(sleep)
sleep_negint <- sleep[,-c(1,2)]
str(sleep_negint)
head(sleep_negint)
```

##  3.  Subsetting in base R - lists and double bracket `[[]]` notation
You can also subset lists. 
```{r, eval=FALSE}
?"[["
```
Create an examlpe list:
```{r}
example_list <- list(TRUE, "string data", 5)
example_list
```
Single brackets `[]` will return the list container as well as its value:
```{r}
example_list[1]
```
However, double brackets will return only the value:
```{r}
example_list[[1]]
```

# **Challenge 1**
1. How many ways can you subset the `iris` dataset using column names and positive and negative integers? Type and run `data(iris)` to load the dataset.  
2. How do you view the structure of `iris`?

# 4.  Missing data (NA)
Identifying missing data can be important for subsetting purposes. R codes missing values as `NA`. Identifying missing data is important because dealing with it might be necessary to run basic tests like `mean()`. 
```{r, eval=FALSE}
?NA
```
```{r}
mean(sleep$NonD)    # This returns NA because R is unsure how to deal with NA cells for the `mean()` computation.
```
However, we can use `na.rm = TRUE` to properly calculate the mean of the NonD column by now excluding the NAs. 
```{r, eval=FALSE}
?mean # Scroll down to `na.rm`
```
```{r}
mean(sleep$NonD, na.rm=TRUE) #Now `mean()` returns the mean!
```
While `na.rm()` nor `str()` will not tell us which data are missing in a convenient way, `is.na()` does. Wrap the name of your data frame in `is.na()` to return logical values. Missing data is coded as `TRUE`, while present data are coded as `FALSE`
```{r, eval=FALSE}
?is.na
is.na(sleep)
```
Data are coded as missing in many different ways besides `NA`, so don't be surprised if you see some other signifier. 

## 4.  Missing data (NA) - recoding missing data
Let's recode NA values in place to say "NONE":
```{r}
sleep[is.na(sleep)] <- "NONE"
```
```{r, eval=FALSE}
sleep
```
They now say "NONE".

However, for R to handle them correctly, we want to recode them to say `NA`. We can do this with a combination of the name of our data set, bracket notation, our relational operator `==` and our old friend the assignment operator `<-` ! 
```{r}
sleep[sleep == "NONE"] <- NA
```
```{r, eval=FALSE}
sleep
```

> NOTE: here `<NA>` and `NA` are synonymous and R will treat them both as missing.  

We can also subset only rows without any missing data using bracket notation. `complete.cases()` will find rows with no missing values.
```{r, eval=FALSE}
?complete.cases
```
```{r}
sleep_no_NA <- sleep[complete.cases(sleep),]  
```
Remember to include the comma here to tell R you want ALL of the columns for these rows :) 
```{r, eval=FALSE}
sleep_no_NA
```
Then, test it to see if it contains missing values. All cells are `FALSE`
```{r,eval=FALSE}
is.na(sleep_no_NA)
```

```{r, eval=FALSE}
# Conversely, we can subset the sleep data to include only rows with missing data by adding the logical bash operator `!` (not). 
?"!"
```
```{r}
sleep_NA <- sleep[!complete.cases(sleep),]
```
```{r, eval=FALSE}
sleep_NA # All rows have at least one cell with missing data
is.na(sleep_NA) # Now we see TRUE values where data is missing
```

`subset` can also be used to subset your data. See `?subset` for more information. For example:
```{r}
subset_1 <- subset(sleep, BrainWgt < 10 & Pred == c(1,5) & Danger == 4)
subset_1

# or
subset_2 <- subset(sleep, BrainWgt < 10 & Pred == c(1,5) & Danger == 4, select=c(BrainWgt, Pred, Danger))
subset_2
```

# **Challenge 2**
1. How many different ways can you subset the `sleep` dataset using logical tests for `NA` data?

# 5.  Merging data
Merging data is useful when we want to combine two different dataframes that share a vector/column. We "merge" by that column. The first two arguments in `merge()` are the names of the two data frames, followed by `by` where we tell which column names we want to match.  

Also note this shorthand way for creating a dataframe. Get your typing fingers ready! For example: 
```{r, eval=FALSE}
?merge #Click the "Merge two data frames" link
```
```{r}
df1 <- data.frame(Name=c("Joe", "Susan", "Jack", "Kelly"),
                  City=c("Berkeley", "Berkeley", "Oakland", "Oakland"),
                  Math=c(42, 48, 50, 46),
                  Reading=c(8, 10, 10, 10))
df1

df2 <- data.frame(Name=c("Joe", "Susan", "Jack", "Kelly"),
                    Science=c(99, 100, 99, 100),
                    Music=c(19, 18, 20, 20),
                    Art=c(20, 20, 19, 18))
df2

df_merge <- merge(df1, df2, by="Name")
df_merge
#What happened here? 
```

## 5.  Merging data - `cbind()` and `rbind()`
Other useful functions include `cbind()` and `rbind()`.  

`cbind()` will bind two data frames by their columns. Let's cbind `df1` and a new dataframe, `df3`:
```{r, eval=FALSE}
?cbind
?rbind # Click the "Combine R Objects by Rows or Columns" link
```
```{r}
df1

df3 <- data.frame(Name=c("Heather", "Billy", "Hector", "Jane"),
                  City=c("San Francisco", "Los Angeles", "San Francisco", "Chico"),
                  Math=c(49, 44),
                  Reading=c(10,10))
df3

df_cbind <- cbind(df1, df3)
df_cbind
```
> NOTE: cbind will not warn you if you have duplicate column names! 

`rbind()` will add more a dataframe as rows to an existing data frame. An example:
```{r}
df1 

dfB <- data.frame(Name=c("James", "Cher", "Elizabeth", "Barack"),
                  City=c("Cleveland", "Memphis", "Detroit", "Chicago"),
                  Math=c(44, 49, 50, 49),
                  Reading=c(9, 9, 9, 10))
df_rbind <- rbind(df1, dfB)
df_rbind
```

> NOTE: for rbind to work, you should have the same column names for both of your data frames! 

# **Challenge 3** 
1. load your `animals` dataset from Day 1 using `read.csv()`.
2. create a subsetted data frame called `cats_dogs` that contains only cats and dogs.
3. create a subsetted data frame that only contains healthy pigs!  
4. load the `mtcars` dataset. How might you find out what it is and how do you load it? 
5. subset `mtcars` in five different ways.  
 
## 6. Reshaping your data
You will probably find that certain plotting and testing functions require that the data be formatted in "long" or "wide" format. The "reshape2" R package offers a handy way to reformat your data. 

In base R, `t` will quickly transpose your dataframe or matrix. The rows will become the columns, and the columns will become the rows:
```{r, eval=FALSE}
?t
```
```{r}
mat1 <- matrix(1:10, nrow=5, ncol=2)
mat1
mat1_t <- t(mat1)
mat1_t

# or

animals_t <- t(animals)
animals_t
```

## 6. Reshaping data with the "reshape2" package 
```{r, eval=FALSE}
library(reshape2)
```

"Wide" format means that each variable is shown in a column, like in our animals dataframe:
```{r}
head(animals)
```

"Long" format means that each row is a unique combination of our "id" variable (animal "Type") and each variable ("Healthy", "Weight", "Height", and "Progress").  

##### `melt` and `cast`
Melt is a handy way to reshape data from "wide" to "long" format:
```{r}
?melt
?dcast
```

```{r}
animals_melt <- melt(animals, id="Type")
str(animals_melt)
```
```{r, eval=FALSE}
animals_melt
```

Cast is handy to perform some basic summary operations. For example, let's look at the mean Weight of each animal Type:
```{r}
type_means <- dcast(animals_melt, Type~variable, mean)
type_means

type_sd <- dcast(animals_melt, Type~variable, sd)
type_sd
```
We now have a summary table in "wide" format!  

See the links below for more information about "reshape2" and check the [D-Lab calendar](http://dlab.berkeley.edu/calendar-node-field-date) for the "Data Wrangling and Manipulation in R" workshop on reshaping data with the "dplyr" R package.  
[Sean C. Anderson's Introduction to the "reshape2" package](http://seananderson.ca/2013/10/19/reshape.html)  
[Hadley Wickhams introduction to the "reshape2" package](http://had.co.nz/reshape/introduction.pdf)  
[Rob Kabakoff's examples of "reshape2" (and also of everything R) are always worthwhile!](http://www.statmethods.net/management/reshape.html)  

# Acknowledgements
[Wickham H, Grolemund G. 2016. R for Data Science](http://r4ds.had.co.nz/)
