---
title: 'R Fundamentals Part 4: For loops and functions'
author: "D-Lab"
date: "September 12, 2017"
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is an R markdown file (.Rmd). Check out (yihui)[https://yihui.name] to learn a little about the creator of `knitr` and a contributor to (`rmarkdown`)[http://rmarkdown.rstudio.com/lesson-1.html].  

Part 1 review:  
1. Variable definition (three part recipe)
2. Data types
3. Data structures

Part 2 review:  
1. Set your working directory to the R-Fundamentals "data" folder.
2. Load the `sleep_VIM` data set. 
3. One (`$`) versus two dimensional (`[ , ]`) subsetting
```{r}
sleep_VIM <- read.csv("./data/sleep_VIM.csv", header=T)
```

Part 3 review:  
1. Summarizing data  
  `summary`, `describe`, `table`
2. Plotting data  
  `hist`, `plot`, `boxplot`
3. ggplot2 - what do you need to use `ggplot`?
  a) a dataset
  b) aesthetics
  c) geoms
4. Statistical testing
  `t.test`, `aov`, `TukeyHSD`, `cor.test`, `lm`
  
# Part 4 Learning objectives  
1. Introducing for-loops and functions
2. For loops
3. Functions
4. Monte Carlo simulation
5. (The birthday problem)[http://mathforum.org/dr.math/faq/faq.birthdayprob.html]

# 1.  Introduction  
Commands such as `mean` are actually functions. Functions in R are a set of instructions that together can provide smart, reusable ways to conduct research. Assign functions in a way similar to how you assign an object (with `<-`). Functions are generally meant to be mapped to data structures or are used to write other functions. Using functions might allow you to write your code more efficiently.  

What do you think the `nrow` function does? Type a function name without the parentheses to see its underlying functional structure. Let's try `nrow`, which returns the number of rows in a data frame. 
```{r, eval=F}
nrow
```
```{r}
# Note that doing this manually or via nrow gives the same result:
dim(sleep_VIM)[1L] == nrow(sleep_VIM)
```

We can do the same with the mean function as well: 
```{r}
mean
(1+2+3)/3 == mean(c(1,2,3))
```

However, before we get to functions, let us review for loops because many functions have for loops embedded in them.   

# 2.  For loops  
Before we introduce how functions work and learn how to write them, let us discuss how a for loop works in R. A for loop repeats a block of code a certain number of times until a certain condition is met (or is no longer met), telling the code to stop and potentially print/return some sort of output. 

The basic syntax looks like this:

syntax: `for(variable in sequence) {statement}`

All for loops are precluded by `for` so that R knows you want to iterate over a loop. 

The `variable` in `(variable in sequence)` is generally denoted with an `i`, which stands for "iterator". However, `i` should be thought of as a placeholder and can be represented by other means (e.g., `x`, `donut`, etc.). `sequence` is some number telling R how many times you want to iterate the code. 

`{statement}` refers to the code that you want run over the sequence at each iteration `i`. Notice that it is contained within curly braces { } - this defines the boundary of the statement in the for loop.  

### 2.1 For-loops - Example 1  
Let's create a matrix of 2 to the power of `i` where `i` is 1 to 10. First we create an NA vector called `vec` which will serve as a placeholder for the output of the loop.
```{r}
vec <- c(rep(NA, 10))
vec
```
Now, let's use a for-loop to create a matrix of 2 to the power of i where i is 1 to 10. 
```{r}
for(i in 1:10){
    vec[i] <- 2^(i)
    }                
vec
class(vec)
```
This code tells R that we want to raise 2 to the power of `i`, where `i` is 1:10.

We can also use indexing to modify only some elements. We now have an object called `vec` with defined values. What if we want to replace the first 5 elements of `vec` with 3 to the power of `i` instead?  

You can change the `sequence` to tell R just to overwrite the first five positions!  
```{r}
for(i in 1:5){
    vec[i] <- 3^i
    }                
vec
#This has changed only the first five entries of `vec`
```

### 2.2 For-loops - Example 2
You can also loop over character data. First, create a name vector:
```{r}
animal.names <- c("Cat", "Dog", "Pig", "Elephant", "Giraffe")
animal.names
```
Then, we create an NA vector of the same length like we did above:
```{r}
animals.length <- rep(NA, length(animal.names))
animals.length
```
Now give the NA vector names:
```{r}
names(animals.length) <- animal.names
animals.length  #See how we are building this from scratch?
```
Finally, perhaps you want to perform some operation across `animals.length`. For example, count the number of characters in each animal name like this:
```{r}
for(i in animal.names){
  animals.length[i] <- nchar(i)
}
animals.length
```

### 2.3 For loops (`break` and `stopifnot`)  
For long loops, many intermediate commands can help us "end the looping" once a condition is met, or when a condition is no longer satisfied. Two particular useful ones are `break` and `stopifnot`.  

`break` ends the looping once a certain condition is met.

`stopifnot` ends the loop when a certain condition is NOT met.  

Let's begin by writing a for loop that outputs the numbers of the sequence 1:100  
```{r, eval=FALSE}
for(i in 1:100){
    print(i) 
    }
```

##### 2.3.1 For-loops (`if`)
`if` statements are frequently used to specify code to be evaluated when the condition is held. Here we tell the code that when `i` is equal to 50, stop the code.  

We usually have data that we pass into a for loop. Let's create an object called `my_nums` that contains the integer vector 40:70. Let's test the _value_ (rather than the index):
```{r}
my_nums <- 40:70
for(i in 1:length(my_nums)){
    print(my_nums[i])
    if(my_nums[i] == 50) break  
}
# Numbers up to and including 50 are printed! 
```

Let's quickly change our `my_nums` object to include one `NA` in position 15: 
```{r}
my_nums[15] <-  as.numeric(NA)
my_nums
```

`stopifnot` works similarly. Here we tell the code to stop if `i` reaches a missing value (`NA`):
```{r, eval=FALSE}
for(i in 1:length(my_nums)){
    print(my_nums[i])  
    stopifnot(!is.na(my_nums[i]))
}
```

#**Challenge 1**
1. Write a for loop that outputs something. 
```{r}

# your code here

```

### 2.4 For loops (`if`, `else`, and `ifelse`)
`if` and `else` statements are control structures that let you control how a code should be iterated within a single for loop. You saw `if` above and is handy when you want to assign different tasks to different subsets of data using a single for loop.  

`if` something happens, do "this",    
if something `else` happens, do "that"!

syntax: `if (condition) {statement} else {other statement}`

```{r, eval=FALSE}
x <- 1
if(x > 7){
  print(x)
}else{   #`else` should not start its own line. Always let it be preceded by a closing brace on the same line.
  print("NOT BIG ENOUGH!!")
}
```
Reassign `x <- 10` here - what happens? Does indentation matter?  
This also works in a loop. Here, we get all outputs for the loop:
```{r}
x <- 1:10
for (i in 1:length(x)){
    if(x[i] > 7){
        print(x[i])
    }else{                     
      print("NOT BIG ENOUGH!!")
    }
}
#Super cool! :) 
```

The `ifelse` function can be handy to recode data as long as you have two conditions that are binary/mutually exclusive. 

syntax: `ifelse(test, yes, no)`

Think back to the random sampling portion of part 1. Generate ten random draws of two common housepets "cat" and "dog":
```{r}
set.seed(1)
animal <- sample(c("cat", "dog"), 10, replace=TRUE)
animal
```
Now we can recode the character data to numeric using `ifelse()`:
```{r}
animal_int <- ifelse(animal == "cat", 1, 0)       
animal_int
```
... or recode the numeric data to some other character type:
```{r}
animal_char <- ifelse(animal_int == 1, "meow!?", "WOOF")
animal_char
```
... and recode the new character data type to logical type:
```{r}
animal_bool <- ifelse(animal_char == "meow!?", TRUE, FALSE)
animal_bool
```

Finally, we can recode it back to its original format:
```{r}
animal2 <- ifelse(animal_bool == TRUE, "cat", "dog")
animal2

animal == animal2 # or
sum(animal != animal2)
```

#**Challenge 2**
1. Write a for loop that uses `if` and `else`
```{r}

# your code here

```

2. Create and then recode some binary data using `ifelse`
```{r}

# your code here

```

# 3.  Functions
In R, a function is a set of code that might be more useful if it is self-contained and/or is to be used repeatedly. For example, if you want to run the same satatistical tests on several datasets, you might want to write a function that contains instructions for all the tests once, so that you do not have to rewrite the code each time.  

Every function has four parts and its basic syntax looks like this:  

> function.name <- function(x){
>   body of function
> }

1. `function.name` - like saving variables in R, it is helpful if you give your function a relevant name.
2. `function(x)` - `function` lets R know you are writing a function and `(x)` contains the inputs/arguments.
3. `{body of function}` - the body of the function is contained within curly braces { } and denotes the statements that you want R to evaluate.
4. The `environment` (global environment) that the function operates within. 

### 3.1 Functions - Example 1
```{r}
test_function <- function(x){
  x + 1
}

class(test_function)        # Returns the class of `test_function`
formals(test_function)      # Shows the defined arguments
body(test_function)         # Displays the statements to be evaluated
environment(test_function)  # Returns the "global" environment
test_function               # Shows your function as you have written it
```

See how our function has `R_GlobalEnv` as it's environment? That's because we defined it in the global environment. This means that if you tell a function to look for an `object`, it will look in the global namespace - not within the function itself.

To evaluate this function, call the name of the function with your argument `x`  within parantheses. R then evaluates the body of the function and returns the desired output for the specified input. 

If we want to see what `x + 1` is equal to when `x` = 2, we type:
```{r}
test_function
test_function(2)
```

### 3.2 Functions - Example 2 - default arguments  
We can also add a predetermined value for one or more arguments, which serves as a default value that we can change in particular applications. You can change the value of "y" - it is an optional argument because it is named and it has a default value. By convenction, required arguments do not have default values and are listed first in the function definition.
```{r}
f <- function(x, y=2){
  (x + y) / y
}
f(5)
f(4, 4) # here, we do not need to specify `x` and `y` because R knows that the second argument is "y"
f(4, y=4) # this is the same as f(4, 4)

#does f(y=4, 4) work? Yes, but it is not recommended.
f(y=4, 4)
```

### 3.3 Functions - Example 3 - function environment
Let's try one more example where we define object `z` within the function. 
```{r}
f <- function(x, y=2){
  z <- 5
  (x + y) / z
}
f(4, 4)

ls()

# the function envrionment can pull objects from the global environmnet
# the gloval environmental CANNOT pull objects from the function envrionment 
```
Now we ask: does `z` appear in your global environment? Why not? `z` is protected from what is happening outside the function!  

Finally, let's think about this a little differently. Now let's define `z` in the global environment and see if the function can access it:
```{r}
z <- 5 # define z outside of the function environment

f <- function(x, y=2){
  (x + y) / z
}
f(4, 4)

ls()
```
`z` is now defined in the global environment and the function can pull objects from the global namespace. For example, using constants that never change, define them in the global namespace. If it is unique to the function however, keep it in the function! Scope is tricky! 

### 3.4 Functions - functions-within-functions
It is also common for functions to be declared within another function.
```{r}
f <- function(x) {
  y <- 1
  g <- function (x) {
    x + y
  }
  g(x)
}
f(1)
```
What is going on here? Does `g` show up in your global environment? Why not? (hint: because it was defined in the functional environment instead of the global environment!)

This is important because it means that functions can be separated from the state of your computer (which is what makes them easy to parallelize). What we really mean here is that anything created inside the function environment doesn't show up in the global environment. 

> side note - R automatically returns the value of the last expression, so there is no need for an explicit `return` statement unless you want to break the function early

For example:
```{r}
f <- function(x) {
    if (x>5) {return("ERROR")}
    y <- 4
    g <- function (x) {
    x + y
  }
  g(x)
}
f(1)
f(10)
```

## 3.  Functions - unit conversion example
Unit conversion is a common obstacle in research. Let us write a function that converts inches to centimeters. For now however, pretend that we think that one inch is equal to 2.5 centimeters - we will show you how to update it below. 

Define a function called `in_to_cm` and then enter Evan's height (74 inches):
```{r}
in_to_cm <- function(x){
  x * 2.5
}
in_to_cm(74)
#Evan is 185 cm tall (incorrectly assuming that 1 inch = 2.5 cm)
```
What if we want to know how tall I am in meters?

You could type:
```{r, eval=FALSE}
function(x){
  x * 2.5 / 100
}
```
...but this would be repeating yourself!

Then, when you figure out that the conversion factor is really *2.54*, not 2.5, you might update one function and forget to update the other - these inconsistencies can cause problems.

Instead, let's define a new function called `in_to_m` so that the output of `in_to_cm` is used in the new function! 
```{r}
in_to_m <- function(x){
  in_to_cm(x) / 100
}
in_to_m(74)
```
Now, if we go back and update `in_to_cm`, those changes automatically get propogated to `in_to_m` and we do not have to worry about updating it: 
```{r}
in_to_cm <- function(x){
  x * 2.54
}
in_to_m(74)
#Evan is actually 1.8796 meters tall
```
R is a bit quirky in that there is no such thing as an uncontained value, e.g. the number `74` is really a vector with length of one, and a value of 74 in position 1
```{r}
74 == c(74)
```
This means that R automatically broadcasts functions across vectors of any length. Notice that there is only one argument ("height.vec") even though it contains four different heights.
```{r}
height_vec <- c(74,64,73,82)
in_to_m(height_vec)
```
Note this doesn't work with lists:
```{r, eval=F}
height_list <- list(74,64,73,82)
in_to_m(height_list)
```

Instead, a functional such as `lapply()` might be used to do so. Remember that a functional is a function that takes functions as arguments. 
```{r, eval=F}
lapply(height_list, in_to_m)
```

Although the "dplyr" R package is perhaps more useful for quickly applying functions to a matrix, there exists a whole family of functionals for this sort of thing in base R. Learn more about the "apply" functional family [here](https://www.r-bloggers.com/using-the-apply-family-of-functions-in-r/), [here](https://www.r-bloggers.com/using-apply-sapply-lapply-in-r/), and [here](http://adv-r.had.co.nz/Functionals.html).  

Also see (this helpful walkthrough)[http://adv-r.had.co.nz/Functions.html].  

# Break

# 4.  Monte Carlo simulations
A Monte Carlo simulation is an algorithm that relies on repeated random sampling to obtain numerical results. In R, loops and functions are key for Monte Carlo simulations.  

See (this example)[https://www.manufacturing.net/article/2012/11/demystifying-monte-carlo] to learn more.  

We can simulate a die roll by taking a random sample from a 1:6 vector.
```{r}
sample(1:6, 1)

# We can also use the sample function to simulate 100 die rolls...
sample(1:6, 100, replace=TRUE)
```
But what if we wanted to repeat the process 200 times and get the mean of the die rolls for each iteration? One option is to do that with a loop:
```{r, eval=FALSE}
iter <- 200
nr_rolls <- 100
for (i in 1:iter){
    rolls <- sample(1:6, nr_rolls, replace=TRUE)
    #print(rolls)
    print(mean(rolls))
}
```
Another alternative is to write a function that produces one iteration of the process and then use the `replicate` command to repeat the process 100 times. 

### 4.1 Monte Carlo simulation - Step 1
Write a function that works through the process once
```{r}
die_roll_mean <- function(nr_rolls){
    rolls <- sample(1:6, nr_rolls, replace=TRUE)
    mean(rolls)
}

die_roll_mean(nr_rolls = 100)
```

### 4.2 Monte Carlo simulation - Step 2
Now we use `replicate` to iterate the function 200 times. However, what happens to the mean variation based on 100 rolls? What about on just five rolls?  
```{r}
# 100 roll mean
reps <- replicate(200, die_roll_mean(nr_rolls = 100)) 
```
```{r, eval=FALSE}
reps
```
```{r}
# 5 roll mean
reps <- replicate(200, die_roll_mean(nr_rolls = 5)) 
```
```{r, eval=FALSE}
reps
```

What happens to the values with more rolls? With less rolls?  

##### 4.2.1 Monte Carlo simulation - include a plotting function
We can use functions with plotting functions as well:
```{r}
hist(reps)

hist(reps, col="grey")
abline(v=mean(reps), col="blue", lwd=6)

my_hist <- function(sims){
    hist(sims, col="grey", xlab="Mean of n die rolls", main="")
    abline(v=mean(sims), col="blue", lwd=6)
}

my_hist(sims=reps)
```

### 4.3 Monte Carlo simulation - Step 3
How would we update the function to add the number of die rolls as the title of the plot?
```{r}
my_hist <- function(sims, n){
    hist(sims, col="grey",  xlab="Mean of n die rolls",
         main=paste0("n=",n), xlim=c(2,5))
    abline(v=mean(sims), col="blue", lwd=2)
    }

my_hist(sims=reps, n=5)
```
How can we use everything we learned today to analyze the changes in the the sampling distribution of the mean of the die rolls as the number of die rolls change?

We can integrate what we did above and write a loop that varies the number of die rolls over which we take the mean and plots the sampling distribution each time.

Say we want to see the sampling distribution for $n=10$, $n=25$, $n=50$, $n=100$, $n=200$, and $n=500$.
```{r}
nr_die_rolls <- c(10, 25, 50, 100, 200, 500)
```
Let's see how we can first write this for $n=10$. Remember this is the first element of the vector, so we can get it using 1 as the index, `nr_die_rolls[1]`.
```{r}
reps <- replicate(1000, die_roll_mean(nr_rolls = nr_die_rolls[1])) 
my_hist(sims=reps, n=nr_die_rolls[1])
```
We can now write a loop that works through the vector with the numnber of die rolls and produces the relevant plot for each iteration:
```{r}
par(mfrow=c(2,3))

for (i in 1:length(nr_die_rolls)){
    reps <- replicate(1000, die_roll_mean(nr_rolls = nr_die_rolls[i]))
    my_hist(sims=reps, n=nr_die_rolls[i])
}
```

#**Challenge 3:** The Birthday Problem
Say we choose 25 people at random. What is the probability two or more of them have the same birthday?'

Model simplifications:
* Ignore leap years
* Assume the probability of having a birthday in any of the 365 days of the year is equally likely (and no twins, triplets, etc. in the room).

Let's start by thinking of one room. We need to sample 25 birthdays and check how many are repeated. 
```{r}
set.seed(94705)  
birthdays <- sample(1:365, 25, replace=TRUE)
birthdays
```
We now want to know how many of those birthdays are repeated. For that we can use the `unique()` command.
```{r}
unique(birthdays)
length(unique(birthdays))
```
24 birthdays are unique. Two people have the same birthday!  

Note this is just one realization of the process. To use R to approximate the probability we would need to repeat this process many, many times. 

For this we can start by writing a function that goes through the process one time and then use the `replicate` function as we did above. 

We start by writing the function:
```{r}
set.seed(94705)
birthday_function <- function(people=25){
    
    # we populate the room
    birthdays <- sample(1:365, people, replace=TRUE)

    # get the unique number of bdays
    unique_bdays <- length(unique(birthdays))

    # and return a 1 if at least one bday is repeated.
    as.numeric(unique_bdays!=people)
}

birthday_function(people=25)
```
Thus, we have one repeat birthday! 

Now we can use `replicate` to repeat the process 1,000 times:
```{r}
set.seed(94705)
many_sims <- replicate(1000, birthday_function(people=25)) 
```
```{r, eval=FALSE}
many_sims
```
To approximate the probability of at least one matching birthday, we can just take the mean of this vector:
```{r}
mean(many_sims)
```
Say we now want to use R to see how this probability changes as the number of people in the room changes. How can we do this? We can use our function but incorporating it into a loop which varies the number of people in the room:
```{r}
people <- 2:100
sims <- matrix(NA, length(people), 2)

for(i in 1:length(people)){
    many_sims <- replicate(5000, birthday_function(people=people[i]))
    sims[i,] <- c(people[i], mean(many_sims))
}

head(sims)
```
```{r, eval=FALSE}
sims
```

Plot the probability: 
```{r}
par(mfrow=c(1,1))
plot(sims[,1], sims[,2], pch=16, col="blue", 
     xlab="Nr of People", ylab="Probability of at Least One Match")
```

Learn more about the birthday problem [here](http://mathforum.org/dr.math/faq/faq.birthdayprob.html)

# Acknowledgements
[Software Carpentry](https://swcarpentry.github.io/)  
[Hadley Wickham](http://adv-r.had.co.nz/)  
[more Hadley Wickham](http://r-pkgs.had.co.nz/)
- Contributions by Guadalupe Tuñón, Dillon Niederhut, Shinhye Choi, Rochelle Terman, Evan Muzzall. 