---
title: 'R Fundamentals Part 2: Subsetting and reshaping'
author: "D-Lab"
date: "2017/09/06"
output:
  html_document:
    toc: yes
    toc_float: yes
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Day 1 review
1. Set your working directory
2. The assignment operator `<-`
3. Atomic data types: numeric, character, logical
4. Data structures: vector, list, matrix, data frame
5. Save your work: `write.csv()`

Part 2 learning objectives:
1. Day 1 review
2. Loading data from files
3. Subsetting in base R
4. Missing data (NA)
5. Merging data
6. Time for questions? / Day 1 and 2 review and extra exercises
7. Exploring data and plotting - preview

# 1.  Loading data from files
# **Challenge 1**
1. Set your working directory to the R-Fundamentals data folder (remember, there are two ways to do this!

Now we can load the animals dataset that you created in Part 1. It is conveniently located in the "data" folder. Note that we can also specify the file path within `read.csv`:
```{r}
?read.csv
```
```{r}
animals <- read.csv("./data/animals.csv", header=TRUE, stringsAsFactors=FALSE)

# . means "current folder"
# .. means "folder above"
```
```{r, eval=F}
# or with "absolute" file path:
animals <- read.csv("/Users/evanmuzzall/Desktop/R-Fundamentals/data/animals.csv", header=TRUE, stringsAsFactors = FALSE) #na.strings = T
```
> Notice that `stringsAsFactors = FALSE`. If set to `TRUE`, R will try to guess which character vectors should automatically be converted to factors. This is problematic because 1) R is not always good at guessing and 2) R defaults to alphabetical and increasing numeric factor level sorting. This might not matter for your data, but it is recommended to set `stringsAsFactors = FALSE` and manually convert your desired character vectors to factors. Refer back to the end of Part 1 for these instructions. `header = TRUE` will include the header row; `header = FALSE` will turn your header row into the first row of actual data. `na.strings = TRUE` allows you to specify data that should be automatically converted to `NA` upon importation. 

# **Challenge 2**
1. Load the mammals "sleep_VIM" dataset from the data folder. 

```{r}
sleep_VIM = read.csv("/Users/evanmuzzall/Desktop/R-Fundamentals/data/sleep_VIM.csv", header = T)
```

This dataset is part of the [VIM package](https://cran.r-project.org/web/packages/VIM/VIM.pdf). It looks at sleep patterns in various mammal species and contains missing (`NA`) values. Click  [here](http://people.stat.sfu.ca/~cschwarz/Stat-650/Notes/MyPrograms/MultReg/AnimalSleep/AllisonScience1976.pdf) to see the original research article. 

> NOTE: you DO NOT want to load the Student's Sleep Dataset from the "datasets" library. 

> Your sleep dataset should have **62 rows and 10 columns**.  

When dealing with Microsoft Excel files (.xlsx), you might find it more convenient to save them first as .csv files in Excel and then import them using `read.csv()`.  

Other functions also work to import data from files, such as `load()`. You might also have success with the "xlsx" R package and its `read.xlsx()` command for directly importing Excel files.  

Also, the "foreign" R package has commands for loading data from SAS, SPSS, Stata, etc.

# 2.  Inspecting the data frame
Remember from Part 1 that we can learn a lot about data in R. For dataframes, the following commands are useful:
```{r, eval=FALSE}
str(sleep_VIM)    # returns the structure of the dataframe
dim(sleep_VIM)    # dataframe dimensions; remember, we want 62 rows and 10 columns! 

rownames(sleep_VIM)   #row names (they have not been named and default to character type)
# for example:
class(rownames(sleep_VIM))

nrow(sleep_VIM)   # number of rows
ncol(sleep_VIM)   # number of columns
unique(sleep_VIM)   # show rows with unique data

head(sleep_VIM, n = 10) # show the first "n" rows
tail(sleep_VIM, n = 4) # show the last "n" rows
```
`names()` and `colnames()` both return column names of the data frame:
```{r}

names(sleep_VIM)  
colnames(sleep_VIM)
```
We can also check which indices are true. Let's convert the "Span" column to a logical vector where missing data is coded as `NA` while present data is marked `TRUE`.
```{r, eval=FALSE}
?as.logical
```
```{r}
new_Span <- as.logical(sleep_VIM$Span)
new_Span
```
Now we can see which data are missing (`NA`) and which cells have data present (`TRUE`).

`which()` will return the rows numbers that have data present:
```{r, eval=FALSE}
?which
```
```{r}
which(new_Span)
```

# 3.  Subsetting in base R
Efficiently subsetting your data will save you time and stress. Fortunately, there are several different ways to subset data in base R.

### 3.1 Subsetting in base R (`$`)
Remember from Part 1 that the dollar sign operator `$` will extract only a single vector/column within the data frame:
```{r, eval=FALSE}
?"$" # Remember to wrap symbols in quotation marks to view their help pages
```
```{r}
sleep_VIM$Dream #Returns only the "Dream" vector from the `sleep_VIM` data frame.
```

# **Challenge 2**
1. Practice extracting a few variables from the `animals` dataframe using the dollar sign `$`
2. Plot them in a histogram using `hist()`
3. Can all data types be visualized with histograms? Why or why not?

### 3.2 Subsetting in two dimensions: rows and columns (`[c(:), c(:)]`)
However, we do not have to specify all rows or columns when subsetting with bracket notation. We can specify dimensions of both rows and columns that we want.

### 3.3 Bracket notation (`[ , ]`)
However, you might find subsetting using **bracket notation** `[ , ]` along with variable names, positive and negative integers, and/or logical values is easier because you can subset multiple elements at once.  

> Inside the brackets, everything before the comma refers to rows. Everything after the comma refers to columns! 

> [rows , columns]

Let's start just with the columns!  

### 3.4 Variable names (`[,c(Variable Names)]`)
You can subset your data by specifying variable names within bracket notation and using the `c()` command to create a column name character vector of names you want to keep.

We can create a new dataframe object `sleep_varnames` that includes only "BodyWgt" and "BrainWgt" variable names from the `sleep` data frame:
```{r, eval=FALSE}
?"["
```
```{r}
sleep_varnames <- sleep_VIM[,c("BodyWgt", "BrainWgt")]
str(sleep_varnames)
head(sleep_varnames)

# compare this to "sleep_VIM", which remains unchanged!
sleep_VIM
```

> Notice that the comma is still included within the bracket notation before the vector of column names. This indicates that we want ALL of the rows corresponding to these two columns. This is the same when we only want to subset rows and include ALL columns (see below).  

##### Remember that in bracket notation `[ , ]` everything **before** the comma refers to rows, and everything **after** the comma refers to columns!

### 3.5 Subsetting with positive integers 
Subsetting by **positive** integers works as well. This will **include** only the specified column indicies - we do not have to type out their names!  

Let's create an object `sleep_posint` that includes only "NonD", "Sleep", and "Exp" columns. 

First use `str()` to see which integer values these columns represent. Because we only want the 3rd, 5th, and 9th columns, we type:
```{r, eval=FALSE}
str(sleep_VIM)
sleep_posint <- sleep_VIM[ ,c(3,5,9)] # why does our vector go after the comma?
str(sleep_posint)
head(sleep_posint)
```

### 3.6 Subsetting with negative integers 
Subsetting by **negative** integers will **exclude** the specified columns. Notice the `-` symbol before `c()` inside our bracket notation.

We can create an object called `sleep_negint` that includes everything **except** columns 1 and 2 ("BodyWgt" and "BrainWgt"):
```{r}
str(sleep_VIM)
sleep_negint_VIM <- sleep_VIM[,-c(1,2)]
str(sleep_negint_VIM)
head(sleep_negint_VIM)
```

### 3.7 Subsetting with logical tests
We can also use logical tests to subset our data. For example, what if we want to include only the rows that have a value of 5 for "Exp"? We can use the relational operator `==`:
```{r, eval=FALSE}
?"=="
```
```{r}
sleep_logical <- sleep_VIM[sleep_VIM$Exp == 5, ] # why is the comma included? 
sleep_logical
```
Only rows with values of 5 for "Exp" are returned! 

What if we want to subset the data so it includes only the rows that have a value of 1 **or** 3 for Exp and 2 for "Danger"? We would use the "and" `&` logical operator:
```{r, eval=FALSE}
?"&"
```
```{r}
sleep_logical2 <- sleep_VIM[sleep_VIM$Exp == c(1,3) & sleep_VIM$Danger == 2, ] # see how we are still just subsetting rows???
sleep_logical2
```
Or, we can subset a dataframe that includes only rows with "Exp" = 2 **or** 3 and only the "BodyWgt", "BrainWgt", and "Exp" columns. 
```{r}
sleep_numeric_character <- sleep_VIM[sleep_VIM$Exp == c(2,3), c("BodyWgt", "BrainWgt", "Exp")]
sleep_numeric_character
```

> NOTE: see how we have the column names listed after the comma? When we only want to call rows, we still must include the comma to tell R that we want ALL of the columns! For example, if we want to specify only rows that have values of 2 and 3 for "Exp" and want all of the columns, we still must include the column inside our bracket notation after the specified rows:

```{r}
sleep_rows_only <- sleep_VIM[sleep_VIM$Exp == 4, ]
sleep_rows_only
```

# 4.  Subsetting lists with double bracket notation (`[[]]`)
You can also subset lists. 
```{r, eval=FALSE}
?"[["
```
Create an examlpe list:
```{r}
example_list <- list(TRUE, "string data", 5)
example_list
```

Single brackets `[]` will return the list container as well as its value. 

Consider the analogy of going through airport security. When you place your shoes in the plastic bin and then place it on the conveyor belt, you have placed two things: 1) your shoes inside the 2) plastic bin. 

Use single brackets `[]` retrieves _the whole thing - both the bin and your shoes._ 
```{r}
example_list[1]
```

However, double brackets go one layer deeper inside the plastic bin and return _only your shoes!_ 
```{r}
example_list[[1]]
```

# **Challenge 3**
1. Load the `iris` dataset!  
2. What is this dataset? How do you find out?  
3. Subset the `iris` dataset in a variety of ways using column names and positive and negative integers.  
4. Subset `iris` using logical tests, `&`, and `|`  

# 5.  Missing data (`NA`)
Identifying missing data can be important for subsetting purposes. R codes missing values as `NA`. Identifying missing data is important because dealing with it might be necessary to run basic tests like `mean()`. 
```{r, eval=FALSE}
?NA
```
```{r}
mean(sleep_VIM$NonD)    # This returns NA because R is unsure how to deal with NA cells for the `mean()` computation.
```
However, we can use `na.rm = TRUE` to properly calculate the mean of the NonD column by now excluding the NAs. 
```{r, eval=FALSE}
?mean # Scroll down to `na.rm`
```
```{r}
mean(sleep_VIM$NonD, na.rm = TRUE) #Now `mean()` returns the mean!
```
While `na.rm()` nor `str()` will not tell us which data are missing in a convenient way, `is.na()` does. Wrap the name of your data frame in `is.na()` to return logical values. Missing data is coded as `TRUE`, while present data are coded as `FALSE`
```{r, eval=FALSE}
?is.na
is.na(sleep_VIM)
```
Data are coded as missing in many different ways besides `NA`, so don't be surprised if you see some other signifier. 

## 4.  Missing data (NA) - recoding missing data
Let's recode NA values in place to say "NONE". First, let's make a copy of `sleep_VIM` so we do not alter the original:
```{r}
s_VIM <- sleep_VIM

s_VIM[is.na(s_VIM)] <- "NONE"
s_VIM
```

They now say `NONE`.

However, for R to handle them correctly, we want to recode them to say `NA`. We can do this with a combination of the name of our data set, bracket notation, our relational operator `==` and our old friend the assignment operator `<-` ! 
```{r}
s_VIM[s_VIM == "NONE"] <- NA
```
```{r, eval=FALSE}
s_VIM
```

> NOTE: here `<NA>` and `NA` are synonymous and R will treat them both as missing.  

We can also subset only rows without any missing data using bracket notation. `complete.cases()` will find rows with no missing values.
```{r, eval=FALSE}
?complete.cases
```
```{r}
s_complete <- sleep_VIM[complete.cases(sleep_VIM),]  
```
Remember to include the comma here to tell R you want ALL of the columns for these rows :) 
```{r, eval=FALSE}
s_complete
```
Then, test it to see if it contains missing values. All cells are `FALSE`
```{r,eval=FALSE}
is.na(s_complete)
sum(is.na(s_complete))
```

```{r, eval=FALSE}
# Conversely, we can subset the sleep data to include only rows with missing data by adding the logical bash operator `!` (not). 
?"!"
```
```{r}
s_NA <- sleep_VIM[!complete.cases(sleep_VIM),]
```
```{r, eval=FALSE}
s_NA # All rows have at least one cell with missing data
sum(is.na(s_NA)) # Now we see TRUE values where data is missing
```

# Subset with `subset`
`subset` can also be used to subset your data. See `?subset` for more information. For example:
```{r}
subset_1 <- subset(sleep_VIM, BrainWgt < 60 & Danger == 4)
subset_1

# or
subset_2 <- subset(sleep_VIM, BrainWgt < 100 | Danger == 4, select=c(BrainWgt, Danger))
subset_2
```

# **Challenge 4**
1. Subset the `animals` dataframe: 
a) in one dimension  
b) in two dimensions  
c) using logical tests

# 6.  Merging data
Merging data is useful when we want to combine two different dataframes that share a vector/column. We "merge" by that column. The first two arguments in `merge()` are the names of the two data frames, followed by `by` where we tell which column names we want to match.  

Also note this shorthand way for creating a dataframe. Get your typing fingers ready! For example: 
```{r, eval=FALSE}
?merge #Click the "Merge two data frames" link
```
```{r}
df1 <- data.frame(Name=c("Joe", "Susan", "Jack", "Kelly"),
                  City=c("Berkeley", "Berkeley", "Oakland", "Oakland"),
                  Math=c(42, 48, 50, 46),
                  Reading=c(8, 10, 10, 10))
df1

df2 <- data.frame(Name=c("Joe", "Susan", "Jack", "Kelly"),
                    Science=c(99, 100, 99, 100),
                    Music=c(19, 18, 20, 20),
                    Art=c(20, 20, 19, 18))
df2

df_merge <- merge(df1, df2, by="Name")
df_merge
#What happened here? 
```

### 6.1  Merging data - `cbind()` and `rbind()`
Other useful functions include `cbind()` and `rbind()`.  

`cbind()` will bind two data frames by their columns. Let's cbind `df1` and a new dataframe, `df3`:
```{r, eval=FALSE}
?cbind
?rbind # Click the "Combine R Objects by Rows or Columns" link
```
```{r}
df1

df3 <- data.frame(Name=c("Heather", "Billy", "Hector", "Jane"),
                  City=c("San Francisco", "Los Angeles", "San Francisco", "Chico"),
                  Math=c(49, 44),
                  Reading=c(10,10))
df3

df_cbind <- cbind(df1, df3)
df_cbind
```
> NOTE: cbind will not warn you if you have duplicate column names! 

`rbind()` will add more rows to an existing data frame. An example:
```{r}
df1 

dfB <- data.frame(Name=c("James", "Cher", "Elizabeth", "Barack"),
                  City=c("Cleveland", "Memphis", "Detroit", "Chicago"),
                  Math=c(44, 49, 50, 49),
                  Reading=c(9, 9, 9, 10))
df_rbind <- rbind(df1, dfB)
df_rbind
```

> NOTE: for rbind to work, you should have the same column names for both of your data frames! 

# **Challenge 5** 
1. Make sure your `animals` dataset is loaded. Remember to use `read.csv()` to import the data from the `data` folder. 
2. Create a subsetted data frame called `cats_dogs` that contains only cats and dogs. (hint: does "and" actually mean "and" here?)  
3. Create a subsetted data frame that only contains healthy pigs! (hint: what is the meaning of "and" here in contrast to question #2)  
4. load the `mtcars` dataset. How might you find out what it is and how do you load it?  
5. subset `mtcars` in five different ways.  

You can also order your data using `order`, for example:
```{r, eval=FALSE}
data(mtcars)
new_cars <- mtcars[order(mtcars$cyl, -mtcars$mpg),] # sort by ascending cyl and descending mpg
new_cars
```
 
# 7. Reshaping your data
You will probably find that certain plotting and testing functions require that the data be formatted in "long" or "wide" format. The "reshape2" R package offers a handy way to reformat your data. 

In base R, `t` will quickly transpose your dataframe or matrix. The rows will become the columns, and the columns will become the rows:
```{r, eval=FALSE}
?t
```
```{r}
mat1 <- matrix(1:10, nrow=5, ncol=2)
mat1
mat1_t <- t(mat1)
mat1_t

# or

animals_t <- t(animals)
animals_t
```

### 7.1 The "reshape2" package 
```{r, eval=FALSE}
library(reshape2)
```

"Wide" format means that each variable is shown in a column, more like in our animals dataframe:
```{r}
head(animals)
```

"Long" format means that each row is a unique combination of our "id" variable (animal "Type") and each variable ("Healthy", "Weight", "Height", and "Progress").  

##### `melt` and `cast`
Melt is a handy way to reshape data from "wide" to "long" format:
```{r, eval=F}
?melt
?dcast
```

```{r}
library(reshape2)
iris_melt <- melt(iris, id = "Species")
str(iris_melt)
```
```{r, eval=FALSE}
iris_melt
```

Cast is handy to perform some basic summary operations. For example, let's look at the mean Weight of each animal Type:
```{r,eval=F}
library(reshape2)
type_means <- dcast(animals_melt, Type~variable, mean)
type_means

type_sd <- dcast(animals_melt, Type~variable, sd)
type_sd
```
We now have summary tables in "wide" format!  

See the links below for more information about "reshape2" and check the [D-Lab calendar](http://dlab.berkeley.edu/calendar-node-field-date) for the "Data Wrangling and Manipulation in R" workshop on reshaping data with the "dplyr" and "tidyr" R packages.  
[Sean C. Anderson's Introduction to the "reshape2" package](http://seananderson.ca/2013/10/19/reshape.html)  
[Hadley Wickhams introduction to the "reshape2" package](http://had.co.nz/reshape/introduction.pdf)  
[Rob Kabakoff's examples of "reshape2" (and also of everything R) are always worthwhile!](http://www.statmethods.net/management/reshape.html)  

# Acknowledgements
[Wickham H, Grolemund G. 2016. R for Data Science](http://r4ds.had.co.nz/)  
- Contributions by Evan Muzzall, Shinhye Choi, Rochelle Terman, Dillon Niederhut, Sam Abdel-Ghaffar