---
title: "R Fundamentals Part 4 Project"
author: "Evan"
date: "9/25/2018"
output: 
  html_document: 
    toc: yes
    toc_float: yes
---

# Example project

You are given a dataset you have never seen before. You want to use RStudio to import the dataset and learn about it a little before you proceed with preprocessing tasks. 

##### 1. Load the `mtcars` dataset.  
```{r}
## YOUR CODE HERE
```

##### 2. What are some useful functions to learn more about the data? 
```{r}
## YOUR CODE HERE
```

##### 3. Might you surmise a relationship about something like engine size and miles per gallon?  

##### 4. Create boxplots for mpg as parsed by cyl. 
```{r}
## YOUR CODE HERE
```

##### 5. What does `cor.test` reveal about mpg and cyl?  
```{r}
## YOUR CODE HERE
```

##### 6. Create a scatterplot of the "mpg" and "hp" variables using `ggplot()`. What happens?  
```{r}
## YOUR CODE HERE
```

##### 7. Can cyl be used to predict mpg in a linear regression model? Why must you learn about the assumptions of a statistical test before applying it? 
```{r}
## YOUR CODE HERE
```
  
# Bioarch dataset
### 1. 
Now that you are warmed up, click the play button of the one-line chunk below to generate the bioarch dataset. 
```{r generate bioarchaeology dataset, eval = F, echo = F}
bioarch = data.frame(
  Site = c(Group = c(rep("Huaca Loro", 8),
          rep("Halstatt", 8),
          rep("Çatalhöyük", 8),
          rep("Odai Yamamoto", 8),
          rep("Shum Laka", 8))),
  Sex_est = c("Male","Male","Female","Male","Female","Female","Male","Female",
          "Male","Female","Male","Female","Male","Male","Female","Male",
          "Male","Male","Female","Male","Female","Female","Male","Female",
          "Male","Male","Female","Female","Male","Female","Female","Male",
          "Female","Female","Female","Female","Female","Female","Female","Female"),
  Health = c("Average","Poor","Excellent","Excellent","Poor","Excellent","Poor","Excellent",
             "Average","Average","Average","Average","Excellent","Average","Average","Excellent",
             "Average","Average","Average","Average","Average","Average","Average","Average",
             "Average","Average","Average","Excellent", "Poor","Average","Excellent", "Poor",
             "Excellent","Excellent","Excellent","Excellent","Excellent","Excellent","Excellent","Excellent"),
  Burial_location = c("Cemetery","Temple","Temple","Temple","Cemetery","Temple","Temple","Cemetery",
                    "Hillside","Beach","Hillside","Hillside","Beach","Hillside","Hillside","Beach",
                    "Shrine","Basket","Basket","Basket","Rockshelter","Basket","Basket","Rockshelter",
                    "Temple","Temple","Beach","Temple","Beach","Beach","Temple","Beach",
                    "Cave","Shrine","Rockshelter","Cave","Cave","Rockshelter","Cave","Cave"),
  Depth = as.integer(c(110,120,120,100,80,30,30,30,
                     40,40,50,60,40,40,40,40,
                     80,80,70,80,60,60,60,70,
                     100,120,90,100,90,90,100,120,
                     20,10,20,10,10,15,15,20)),
  Burial_items = as.integer(c(10,8,7,4,6,2,1,1,
                            0,0,1,2,5,8,9,3,
                            3,2,4,4,3,4,3,2,
                            15,18,12,12,14,2,15,18,
                            1,2,0,0,0,0,0,0)),
  Ceramic_vessels = as.integer(c(10,11,14,15,13,1,1,0,
                               3,4,2,2,9,19,1,0,
                               2,3,2,5,6,8,9,3,
                               1,1,1,1,3,0,0,0,
                               1,1,0,0,0,0,0,0)),
  Femur_length = c(480.1,481.0,400.3,482.9,405.7,390.4,380.3,462.5,
                 490.4,440.1,499.8,442.5,502.1,440.0,430.0,490.0,
                 460.5,469.2,455.4,462.1,459.0,460.1,460.2,460.5,
                 480.6,479.9,420.4,422.9,478.7,491.6,490.0,420.0,
                 470.1,469.2,473.4,478.7,471.0,472.4,469.6,419.2))
                  
str(bioarch)
bioarch
```

### 2. 
How would you learn more about this `bioarch` dataset? Where are these sites located? What do the real-life data say about these sites? Can we compare our toy data somehow? What types of questions can we use data to investigate? 
```{r}
## YOUR CODE HERE
```

### 3. 
Pretend you are a bioarchaeologist for the next two hours. You hope to follow the research framework below:

### A) Although you have never seen these data before or know anything about [Bioarchaeology](https://en.wikipedia.org/wiki/Bioarchaeology), write two sentences describing this dataset and write one sentence that proposes one research question to be investigated. 
```{r}
##
##
##
```

### B) Compute descriptive statistics for each variable in the dataset.  
```{r}
## YOUR CODE HERE
```

### C) Can you visualize the data from every single column in this dataset? Create a histogram, boxplot, or scatterplot for the ones you can, and explain in writing the ones you cannot.  
```{r}
## YOUR CODE HERE
```

### D) Do these plots inform you about anything specific that fits your hypothesis and that you can justifiably test? 
```{r}
## 
##
##
```

### E) How might you try to interpret your plots and statistical output? Why? 
```{r}
##
##
##
```

