---
title: "R-Fundamentals solutions"
author: "Evan"
date: "September 5, 2018"
output: html_document
---

# Part 1

### Part 1 - Challenge 1
1. What is the three-piece recipe for variable definition?
2. Define two numeric objects.
3. Define two character objects.
4. What does the `ls()` function do? Where else do you see this information?
5. Remove an individual object using `rm`.
6. Wipe your environment using `rm(list=ls())` or the broom icon. 
```{r}
## YOUR CODE HERE
```

### Part 1 - Challenge 2
1. Define a numeric object as 0 and check its class.  
2. Define a boolean object as "FALSE" and check its class.  
3. Use `==` to compare your numeric and boolean object. Are they equivalent? Why or why not?  
4. Define a character object that uses `paste()` to combine more than one word into a sentence.  
5. Use `substr` to extract the first word of your sentence.  
6. Use `gsub` to replace a character in your sentence with some other character.  
```{r}
## YOUR CODE HERE
```

### Part 1 - Challenge 3
1. Create a character object and check its type using `is.character` and `class`. What is the difference between these two methods?    
2. Try to change ("coerce") this object to another data type using `as.integer`, `as.numeric`, or `as.logical`.  
3. Create an object of class "integer". Remember, there are two ways to do this!  
4. Why is 1 == "1" true?   Why is -1 < FALSE true?   Why is "one" < 2 false?  
```{r}
## YOUR CODE HERE
```

### Part 1 - Challenge 4
Vectors  
1. Create a vector of a sequence of numbers between 1 to 10.  
2. Coerce that vector into a character vector  
3. Add the element "11" to the end of the vector  
4. Evaluate the `str` of the vector.  
```{r}
## YOUR CODE HERE
```

Lists  
5. How does a list differ from an atomic vector?  
6. Create three objects of different types and lengths and then combine them into a list names `x`.  
7. If `x` is a list, what is the class of `x[1]`? How about `x[[1]]`? (this is a preview of Part 2).  
```{r}
## YOUR CODE HERE
```

Data frames  
8. Create a 3x2 data frame called `basket`. List the name of each fruit in the first column and its price in teh second column.  
9. Now give your dataframe appropriate column and row names.  
10. We can add a new column using `$` (this will be covered with data subsetting in Part 2). Can you guess how to add a third column called "Color", that shows the color of each fruit?  
```{r}
## YOUR CODE HERE
```
Basket example solution:  
```{r}
f <- factor(c("Apple", "Orange", "Pear"))
p <- c(10, 28, 36)
basket <- data.frame(f, p)
basket

colnames(basket) <- c("Fruit", "Price")
basket

basket$Color <- factor(c("Red", "Orange", "Green"))
str(basket)
basket
```

### Part 1 - Challenge 5
```{r}
# install.packages("swirl")
library(swirl)
swirl()
```

# Part 2

### Part 2 - Challenge 1
1. Load the mammals "sleep_VIM" dataset from the data folder using `read.csv` (what is shown is another way to load datasets that are included in R and packages). Save it in a variable named "Sleep_VIM"
2. Remember what `ls()` does? What does `dir()` do? 
```{r}
# 1
sleep_VIM = read.csv("sleep_VIM.csv", header = T, stringsAsFactors = F)

# 2
# ls() lists variables defined in the global environment; dir() lists the contents of yoru working directory
```

### Part 2 - Challenge 2
1. Practice extracting two different variables (columns) from the `animals` dataframe using the dollar sign `$`
2. Plot them in a histogram using the `hist()` function. How do you do this? 
3. Can all data types be visualized with histograms? Why or why not?
```{r}

```

### Part 2 - Challenge 3
1. Load the `iris` dataset! Type `data(iris)` to load it.  
2. What is this dataset? How do you find out?  
3. Subset the `iris` dataset in three ways using column names and positive and negative integers.  
4. Subset `iris` using logical tests, `&`, and `|`    
```{r}
## YOUR CODE HERE
```

### Part 2 - Challenge 4
Subset the `animals` dataframe:  
1. in one dimension.  
2. in two dimensions using either bracket notation `[ , ]` or `subset`.  
3. using logical tests.  
```{r}
## YOUR CODE HERE
```

### Part 2 - Challenge 5
1. Using the "animals" data frame, create a subsetted data frame called `cats_dogs` that contains only cats and dogs. (hint: does "and" actually mean "and" here?)  
2. Create a subsetted data frame that only contains healthy pigs! (hint: what is the meaning of "and" here in contrast to question #1?)  
3. load the `mtcars` dataset. How might you find out what it is and how do you load it?  
4. subset `mtcars` in five different ways.  
```{r}
## YOUR CODE HERE
```

# Part 3

### Part 3 - Challenge 1
Use `summary()`, `table()` and `describeBy()` to investigate variables in the `iris` dataset. Create a histogram of one of the numeric variables.  
```{r}
## YOUR CODE HERE
```

### Part 3 - Challenge 2
Create boxplots for one of the numeric variables in the `iris` dataset and export it to your working directory as a .TIFF file. 
```{r}
## YOUR CODE HERE
```

### Part 3 - Challenge 3
Using the `iris` dataset, plot two numeric variables and export the graph to your working directory as a .PNG file. 
```{r}
## YOUR CODE HERE
```

### Part 3 - Challenge 4
In lines 374 - 377 in the chunk above, what are the arguments inside of our second "theme" argument doing?
```{r}
## YOUR CODE HERE
```

### Part 3 - Challenge 5
Subsetting review! Subset your "iris" data frame into three new dataframes: setosa, versicolor, and virginica that each contain only their respective Species. 
```{r}
setosa <- iris[iris$Species == "setosa",]
versicolor <- iris[iris$Species == "versicolor",]
virginica <- iris[iris$Species == "virginica",]

setosa
versicolor
virginica
```

### Part 3 - Challenge 6
Create two numeric vectors. Perform a t.test on them:
```{r}
## YOUR CODE HERE
```

### Part 3 - Challenge 7
Using the `mtcars` dataset, perform an ANOVA and TukeyHSD test for one of the numeric variables between the three species. In general, what are differences between t-test and ANOVA that you should keep in mind?  
```{r}
## YOUR CODE HERE
```

### Part 3 - Challenge 8
1. Load the `mtcars` dataset.  
2. When you load new data into R, what functions are good ones to run to learn more about the data?  
3. Create boxplots for mpg as parsed by cyl. 
4. Might you surmise a relationship about something like engine size and miles per gallon?  
5. What does `cor.test` reveal about mpg and cyl?  
6. Create a scatterplot of the "mpg" and "hp" variables using `ggplot()`. What happens?  
7. Can cyl be used to predict mpg in a linear regression model?  
```{r}
## YOUR CODE HERE
```

# Part 4

### Part 4 - Challenge 1
1. Write a for loop that outputs something. 
```{r}
for (i in 1:10) {
  print(i / 12)
}

# or

for (time in c(12, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11)) {
  print(paste("The time is", time, "AM"))
}
```

### Part 4 - Challenge 2
```{r}
for (x in c(2,4,6,8,10,12)) {
  if (x < 8) {
    print(paste("Whoa the number is", x))
  }else{
    print("nope...")
  }
}

# or

for (i in 1:12) # what happened to the curly braces? 
    if (i < 10) {
        print(paste("The time is", i, "o'clock"))
        }else{
        print("It is getting late!")
}
```

### Part 4 - Challenge 3
Say we choose 25 people at random. What is the probability two or more of them have the same birthday?'
```{r}
## See the Part 4 .Rmd file for the answer
```

