---
title: "R-Fundamentals solutions"
author: "Evan"
date: "September 5, 2018"
output: html_document
---

> Remember to click "Code" --> "Show document outline" to see a clickable list of the answers.  

# Part 1

### Part 1 - Challenge 1
1. What is the three-piece recipe for variable definition? What piece goes where?  
2. Define two numeric variables.  
3. Define two character variables.  
4. How do you check what types of data these variables are?  
5. What does the `ls()` function do? Where else do you see this information?  
6. Remove one of your numeric and one of your character variables.  
7. Try to subract your remaining character. What happens? What might this tell you about data of different types?  
8. Wipe your environment clean.  
```{r}
# 1. 
# A) Unique name does on left side
# B) Assignment operator <- goes in the middle (a single equals sign works just as well)
# C) Definition/statement/expression/ (aka some sort of "code to be evaluated") goes on right side

# 2. 
x = 15
y = 88

# 3. 
my_name = "Oski"
my_city = "Berkeley, California"

# 4.
class(x)
class(y)
class(my_name)
class(my_city)

# 5. 
# ls() will list the variables defined in the global environment (upper right pane in RStudio)

# 6.
rm(x)
rm(my_city)

# 7. 
y - my_name
# we receive an error message! data type is important because it determines what we can do to that specific variable, or group of variables. here, we cannot subtract a string from a numeric variable. 

# 8. 
rm(list = ls()) # or simply click the broom icon in the upper-right RStudio pane
```

### Part 1 - Challenge 2
1. Define a numeric object as 0 and check its class.  
2. Define a boolean object as "FALSE" and check its class.  
3. Use `==` to compare your numeric and boolean object. Are they equivalent? Why or why not?  
4. Define a character object that uses `paste()` to combine more than one word into a sentence.  
5. Use `substr` to extract the first word of your sentence.  
6. Use `gsub` to replace a character in your sentence with some other character.  
```{r}
# 1. 
num_object = 0
class(num_object)

# 2. 
bool_object = FALSE
class(bool_object)

# 3. 
num_object == bool_object # remember FALSE is stored internally as 0, and TRUE as 1

# 4. 
sentence = paste("This", "is", "how", "to", "use", "paste!")
sentence

# 5.
first_word = substr(sentence, 1, 4)
first_word

# 6. 
new_sentence = gsub("to", "we", sentence)
new_sentence
```

### Part 1 - Challenge 3
1. Create a character object and check its type using `is.character` and `class`. What is the difference between these two methods?    
2. Try to change ("coerce") this object to another data type using `as.integer`, `as.numeric`, or `as.logical`.  
3. Create an object of class "integer". Remember, there are two ways to do this!  
4. Why is 1 == "1" true?   Why is -1 < FALSE true?   Why is "one" < 2 false?  
```{r}
# 1.
char_var = "This is character aka string aka text data"
is.character(char_var) # returns logical TRUE or FALSE value
class(char_var) # returns the actual data type

# 2. 
char_int = as.integer(char_var) # error
char_num = as.numeric(char_var) # error
char_logical = as.logical(char_var) # "soft fail!"
char_logical

# 3. 
int1 = as.integer(12)
class(int1)

int2 = 5L
class(int2)

# 4.
1 == "1" # R can see the number inside of the quotations

-1 < FALSE # Remember FALSE is stored internally as 0, TRUE as 1

"one" < 2 # Cannot see if a word is less than a number
```

### Part 1 - Challenge 4
Vectors  
1. Create a vector of a sequence of numbers between 1 to 10.  
2. Coerce that vector into a character vector  
3. Add the element "11" to the end of the vector  
4. Evaluate the `str` of the vector.  
```{r}
# 1.
vector_1 = 1:10
#or
vector_1 = c(1,2,3,4,5,6,7,8,9,10)

# 2.
vec1_char = as.character(vector_1)
vec1_char
class(vec1_char)

# 3.
vec1_more = c(vec1_char, "11")
vec1_more

# 4.
str(vec1_more)
```

Lists  
5. How does a list differ from an atomic vector?  
6. Create three objects of different types and lengths and then combine them into a list names `x`.  
7. If `x` is a list, what is the class of `x[1]`? How about `x[[1]]`? (this is a preview of Part 2).  
```{r}
# 5. 
# Lists are not additive or duplicative like an atomic vector; lists contain a placeholder and a value.

# 6.
obj1 = c(1,4,5)
obj2 = c(T, F)
obj3 = c(15L, 22L, 98L, 10050L)

x = list(obj1, obj2, obj3)
x
class(x)

# 7.
class(x[1])
class(x[[1]])
```

Data frames  
8. Create a 3x2 data frame called `basket`. List the name of each fruit in the first column and its price in teh second column.  
9. Now give your dataframe appropriate column and row names.  
10. We can add a new column using `$`. Can you guess how to add a third column called "Color", that shows the color of each fruit?  
```{r}
# 8. 
f = factor(c("Apple", "Orange", "Pear"))
p = c(10, 28, 36)
basket = data.frame(f, p)
basket

# 9. 
colnames(basket) = c("Fruit", "Price")
basket

rownames(basket) = c("Berkeley Bowl", "Mi Tierra", "Target")
basket

# 10. 
basket$Color <- factor(c("Red", "Orange", "Green"))
basket
str(basket)
```

### Part 1 - Challenge 5
```{r}
# install.packages("swirl")
library(swirl)
swirl()
```

# Part 2

### Part 2 - Challenge 1
1. Load the mammals "sleep_VIM" dataset from the data folder using `read.csv` (what is shown is another way to load datasets that are included in R and packages). Save it in a variable named "Sleep_VIM"
2. Remember what `ls()` does? What does `dir()` do? 
```{r}
# 1
sleep_VIM = read.csv("sleep_VIM.csv", header = T, stringsAsFactors = F)

# 2
# ls() lists variables defined in the global environment; dir() lists the contents of yoru working directory
```

### Part 2 - Challenge 2
1. Practice extracting two different variables (columns) from the `animals` dataframe using the dollar sign `$`
2. Plot them in a histogram using the `hist()` function. How do you do this? 
3. Can all data types be visualized with histograms? Why or why not?
```{r}
# 1.
animals$Type
# or
height = animals$Height

# 2.
hist(height)

# 3.
# No, histograms are used to visualize distributions of one numeric/integer variable.
```

### Part 2 - Challenge 3
1. Load the `iris` dataset! Type `data(iris)` to load it.  
2. What is this dataset? How do you find out?  
3. Subset the `iris` dataset in three ways using column names and positive and negative integers.  
4. Subset `iris` using logical tests, `&`, and `|`    
```{r}
# 1.
data(iris)

# 2. 
str(iris)
head(iris)
?iris
# Google! 

# 3.
iris_colnames = iris[, c("Sepal.Length", "Species")]
head(iris_colnames)

iris_posint = iris[, c(1,5)]
head(iris_posint)

iris_negint = iris[, -c(2,3,4)]
head(iris_negint)

# 4. 
iris_setosa_only = iris[iris$Species == "setosa", ]
table(iris_setosa_only$Species)

iris_setosa_and = iris[iris$Species == "setosa" & iris$Sepal.Length > 5, ]
iris_setosa_and

iris_setosa_or = iris[iris$Species == "setosa" | iris$Species == "versicolor", ]
table(iris_setosa_or$Species)
```

### Part 2 - Challenge 4
Subset the `animals` dataframe:  
1. in one dimension.  
2. in two dimensions using either bracket notation `[ , ]` or `subset`.  
3. using logical tests.  
```{r}
# 1.
one_dim = animals$Healthy
one_dim

# 2. 
two_dim_bracket = animals[1:5 , 1:2]
two_dim_bracket

two_dim_subset = subset(animals, select = c("Type", "Healthy"))
two_dim_subset
```

### Part 2 - Challenge 5
1. Using the "animals" data frame, create a subsetted data frame called `cats_dogs` that contains only cats and dogs. (hint: does "and" actually mean "and" here?)  
2. Create a subsetted data frame that only contains healthy pigs! (hint: what is the meaning of "and" here in contrast to question #1?)  
3. load the `mtcars` dataset. How might you find out what it is and how do you load it?  
4. subset `mtcars` in five different ways.  
```{r}
# 1.
cats_dogs = animals[animals$Type == "Cat" | animals$Type == "Dog", ] 
cats_dogs

# 2.
healthy_pigs = animals[animals$Type == "Pig" & animals$Healthy == TRUE, ]
healthy_pigs

# 3.
data(mtcars)
str(mtcars)
?mtcars
# Google!

# 4. 
# six cylinder cars only
(mt_sub1 = mtcars[mtcars$cyl == 6, ])

# disp greater than 400
(mt_sub2 = mtcars[mtcars$disp > 400, ])

# 5 speed gear and 8 barrel carburetor
(mt_sub3 = mtcars[mtcars$gear == 5 & mtcars$carb == 8, ])

# disp < 100 or hp < 100
(mt_sub4 = mtcars[mtcars$disp < 100 | mtcars$hp < 100, ])

# mpg > 33
(mt_sub5 = subset(mtcars, mpg > 33))
```

# Part 3

### Part 3 - Challenge 1
Use `summary()`, `table()` and `describeBy()` to investigate variables in the `iris` dataset. Create a histogram of one of the numeric variables.  
```{r}
library(psych)
data(iris)
summary(iris)
table(iris$Species)
describe(iris)
describeBy(iris, group = iris$Species)
```

### Part 3 - Challenge 2
Create boxplots for one of the numeric variables in the `iris` dataset and export it to your working directory as a .TIFF file. 
```{r}
boxplot(x = iris$Sepal.Length,
        col = "orange",
        main = "Title goes here",
        xlab = "Sepal Length",
        ylab = "cm",
        las = 1)
# click "Export" 
# click "Save as .PDF"
# open this .PDF
# click "File" --> "Export"
# click "TIFF" with compression type "JPEG"
```

### Part 3 - Challenge 3
Using the `iris` dataset, plot two numeric variables and export the graph to your working directory as a .PNG file. 
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width,
     col = as.numeric(iris$Species),
     pch = as.numeric(iris$Species))

# better way
library(ggplot2)
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, 
                 color = Species, shape = Species)) + geom_point(size = 5, alpha = 0.5) + 
  ggtitle("Title goes here") + xlab("Sepal Length (cm)") + ylab("Sepal Width (cm)") + 
  theme_bw()
```

### Part 3 - Challenge 4
In lines the ggplot code above, what are the arguments inside of our second "theme" argument doing?
```{r}
# change the size of the plot title and justification
# change the legend title size
# change the legend text size
# change the orientaiton and justification of the x-axis text
```

### Part 3 - Challenge 5
Subsetting review! Subset your "iris" data frame into three new dataframes: setosa, versicolor, and virginica that each contain only their respective Species. 
```{r}
setosa <- iris[iris$Species == "setosa",]
versicolor <- iris[iris$Species == "versicolor",]
virginica <- iris[iris$Species == "virginica",]

setosa
versicolor
virginica
```

### Part 3 - Challenge 6
Create two numeric vectors. Perform a t.test on them:
```{r}
a = c(10, 20, 30, 40, 50, 140)
b = c(1, 1.1, 1.4, 1.9, 2.1)
t.test(a,b)
```

### Part 3 - Challenge 7
Using the `mtcars` dataset, perform an ANOVA and TukeyHSD test for one of the numeric variables between the cyl sizes. In general, what are differences between t-test and ANOVA that you should keep in mind?  
```{r}
data(mtcars)
mt_aov = aov(mpg ~ as.factor(cyl), data = mtcars)
summary(mt_aov)
TukeyHSD(mt_aov)
```

### Part 3 - Challenge 8
1. Load the `mtcars` dataset.  
2. When you load new data into R, what functions are good ones to run to learn more about the data?  
3. Create boxplots for mpg as parsed by cyl. 
4. Might you surmise a relationship about something like engine size and miles per gallon?  
5. What does `cor.test` reveal about mpg and cyl?  
6. Create a scatterplot of the "mpg" and "hp" variables using `ggplot()`. What happens?  
7. Can cyl be used to predict mpg in a linear regression model?  
```{r}
# 1.
data(mtcars)

# 2.
str(mtcars)
?mtcars

# 3.
boxplot(mpg ~ as.factor(cyl), data = mtcars,
        col = c("goldenrod", "salmon", "turquoise"),
        main = "mpg verus cyl",
        xlab = "Number of cylinders",
        ylab = "mpg",
        las = 1)

# or
ggplot(mtcars, aes(x = as.factor(cyl), y = mpg, fill = as.factor(cyl))) + 
  geom_boxplot() + 
  theme_minimal() + 
  guides(fill = F)

# 4.
# Yes! The larger the engine, the fewer miles per gallon... (?)

# 5.
cor.test(mtcars$mpg, mtcars$cyl)
# they are strongly and significantly negatively correlated! 
# that is, the more cyl, the less mpg

# 6.
ggplot(mtcars, aes(x = hp, y = mpg, 
                   color = as.factor(cyl), shape = as.factor(cyl))) + 
  geom_point(size = 5, alpha = 0.5) +
  theme_classic()

# or
ggplot(mtcars, aes(x = hp, y = mpg, color = cyl)) + 
  geom_point(size = 5, alpha = 0.5) +
  theme_light()

# 7.
mt_model = lm(mpg ~ cyl, data = mtcars)
summary(mt_model)
```

# Part 4
# Example project

You are given a dataset you have never seen before. You want to use RStudio to import the dataset and learn about it a little before you proceed with preprocessing tasks. 

##### 1. Load the `mtcars` dataset.  
##### 2. What are some useful functions to learn more about the data? 
##### 3. Might you surmise a relationship about something like engine size and miles per gallon?  
##### 4. Create boxplots for mpg as parsed by cyl. 
##### 5. What does `cor.test` reveal about mpg and cyl? 
##### 6. Create a scatterplot of the "mpg" and "hp" variables using `ggplot()`. What happens?  
##### 7. Can cyl be used to predict mpg in a linear regression model? Why must you learn about the assumptions of a statistical test before applying it? 

```{r}
# 1. 
data()
data(mtcars)

# 2. 
?mtcars
str(mtcars)
head(mtcars)
mtcars
View(mtcars)

# field notes, lab reports, scientific instruments, Google, etc. 

# 3. 
# Larger engine equals less miles per gallon? 

# 4. 
boxplot(mpg ~ cyl, data = mtcars,
        col = c("red", "green", "gold"),
        main = "mtcars boxplots",
        xlab = "# of cylinders",
        ylab = "mpg", las = 1)

# or

plot_box = ggplot(mtcars, aes(cyl, mpg, 
                              fill = as.factor(cyl))) + 
  geom_boxplot() + 
  theme(legend.position = "top") + 
  labs(fill = "Cylinders")
plot_box

# 5. 
cor.test(mtcars$mpg, mtcars$cyl)

# 6. 
plot_scatter = ggplot(mtcars, aes(hp, mpg, color = mpg)) + 
  geom_point(size = 8, alpha = 0.8) +
  theme_minimal() 
plot_scatter

# 7. 
lm1 = lm(mpg ~ cyl, data = mtcars)
lm1
summary(lm1)
```
  
# Bioarch dataset
(in class example)

# Bonus 
### Challenge 1
1. Write a for loop that outputs something. 
```{r}
for (i in 1:10) {
  print(i / 12)
}

# or

for (time in c(12, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11)) {
  print(paste("The time is", time, "AM"))
}
```

### Challenge 2
```{r}
for (x in c(2,4,6,8,10,12)) {
  if (x < 8) {
    print(paste("Whoa the number is", x))
  }else{
    print("nope...")
  }
}

# or

for (i in 1:12) # what happened to the curly braces? 
    if (i < 10) {
        print(paste("The time is", i, "o'clock"))
        }else{
        print("It is getting late!")
}
```

### Challenge 3
Say we choose 25 people at random. What is the probability two or more of them have the same birthday?'
```{r}
## See the Part 4 .Rmd file for the answer
```

