---
title: 'R Fundamentals Part 2-2: subset in one dimension'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# Inspecting the data frame

Remember from Part 1 that we can generally learn a lot about data in R. For dataframes specifically, the following functions are useful:

```{r, eval=FALSE}
# Returns the structure/critical information of the data object
str(sleep_VIM)    

# Dataframe dimensions. Rows are the first number, columns are the second. 
# Remember, your dataframe should have 62 rows and 10 columns! 
dim(sleep_VIM)    

# Row names (they have not been named and default to character type)
rownames(sleep_VIM)   
class(rownames(sleep_VIM)) 

# Number of rows
nrow(sleep_VIM)   

# Number of columns
ncol(sleep_VIM) 

# Show rows with unique data
unique(sleep_VIM)   

# Show the first "n" rows (default is 6)
head(sleep_VIM, n = 2) 

# Show the last "n" rows
tail(sleep_VIM, n = 3) 

# `names()` and `colnames()` both return column names of the data frame:
names(sleep_VIM)  
colnames(sleep_VIM)
```

# Subsetting in base R
Efficiently subsetting your data will save you time and stress. Fortunately, there are many different ways to subset data in base R.

## Subsetting in one dimension (`$`)

Remember from Part 1 that the dollar sign operator `$` will extract only a single vector/column within the data frame:

```{r, eval=FALSE}
?"$" # Remember to wrap symbols in quotation marks to view their help pages
```
```{r}
sleep_VIM = read.csv("sleep_VIM.csv")
sleep_VIM$Dream #Returns only the "Dream" vector from the `sleep_VIM` data frame.
```

> Type sleep_VIM$ and press the TAB key - a handy list of the columns in your data will appear! 

```{r, eval = F}
sleep_VIM$
```

# **Challenge 2**

1. Practice extracting a few variables from the `gap` dataframe using the dollar sign `$`
2. Plot them in a histogram using the `hist()` function. How do you do this? 
3. Can all data types be visualized with histograms? Why or why not?

```{r}
## YOUR CODE HERE
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>