---
title: 'R Fundamentals Part2-5: missing data'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# Missing data (`NA`)

Identifying missing data can be important for subsetting purposes. R codes missing values as `NA`. Identifying missing data is important because dealing with it might be necessary to run basic tests like `mean()`. 

```{r, eval=FALSE}
?NA
?mean # Scroll down to `na.rm`
```
```{r}
# This returns NA because R is unsure how to deal with NA cells for the `mean` computation.
mean(sleep_VIM$NonD) 
```

Look at the "Usage" portion of the help file - we know what `rm` is, and we now know what `NA` is, so what do you think the logical argument `na.rm =` will do? That's right! We can use `na.rm = TRUE` to properly calculate the mean of the NonD column by first excluding the NAs. 

```{r}
mean(sleep_VIM$NonD, na.rm = TRUE) #`mean()` now returns the mean!
```

## `is.na()` to identify missing values

While `na.rm()` nor `str()` will not tell us which data are missing in a convenient way, `is.na()` does. Wrap the name of your data frame in `is.na()` to return logical values. Missing data are coded as `TRUE`, while present data are coded as `FALSE`

```{r eval = F}
?is.na
```
```{r}
is.na(sleep_VIM)

# Count the number of total missing cells (total number of TRUE)
sum(is.na(sleep_VIM))
```

People code data in very strange ways besides `NA`, so don't be surprised if you see some other signifier. But, whatever you do, do not use the number zero! 

## Recoding missing data

> Remember that the `na.strings = ` argument inside of the `read.csv()` function is the most convenient way to automatically recode missing data! See Part2-1_load data from fies to review. 

Make a copy of `sleep_VIM` named `sv` so we do not alter the original, then recode NA values in place to say "NONE":

```{r}
sv = sleep_VIM

sv[is.na(sv)] = "NONE"
sv
```

They now say `NONE`.

However, for R to handle them correctly, we want to recode them to say `NA`. We can do this with a combination of the name of our data set, bracket notation, our relational operator `==` and the single equals sign `=` for assigmnet. 

Flash question: can we still use the `is.na` function? Why or why not?  
```{r}
sv[sv == "NONE"] = NA
sv
```

> NOTE: here `<NA>` and `NA` are virtually synonymous and R will treat them both as missing. `<NA>` suggests that for a vector to truly remain a vector, even missing data must be of that vector's type! 

Flash question: uh-oh! run `str(sv)` - what happened to our data types of the recoded columns???

We can also subset only rows without any missing data using bracket notation. `complete.cases()` will find rows with no missing values.

```{r}
?complete.cases
```
```{r}
s_complete = sleep_VIM[complete.cases(sleep_VIM) , ]  
s_complete
```

Remember to include the comma here to tell R you want ALL of the columns for these rows. Then, test it to see if it contains missing values. All cells are `FALSE`

```{r}
is.na(s_complete)
sum(is.na(s_complete)) # what did we do here?
```

Conversely, we can subset the sleep_VIM data to include only rows with missing data by adding the logical bash operator `!` (not). 

```{r, eval=FALSE}
?"!"
```
```{r}
s_NA = sleep_VIM[!complete.cases(sleep_VIM),]
```
```{r}
s_NA # All rows have at least one cell with missing data
is.na(s_NA) # We see TRUE values where data is missing
sum(is.na(s_NA)) # 38 cells have missing data

# Do the number of missing cases in s_NA equal that to the whole data set? 
sum(is.na(s_NA)) == sum(is.na(sleep_VIM)) 
# ... yes! Both data objects contain 38 NA values! 
```
