---
title: 'R Fundamentals Part2-6: merging data'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# Merging data with `merge`

Merging data is useful when we want to combine two different dataframes that share a vector/column. We "merge" by that column ("unique identifier"). The first two arguments in `merge()` are the names of the two data frames, followed by `by` where we tell which column names we want to match.  

First, note this shorthand way for creating a dataframe. For example: 

```{r}
df1 = data.frame(Name = c("Joe", "Susan", "Jack", "Kelly"),
                  City = c("Berkeley", "Berkeley", "Oakland", "Oakland"),
                  Math = c(42, 48, 50, 46),
                  Reading = c(8, 10, 10, 10))
df1

df2 = data.frame(Name = c("Joe", "Susan", "Jack", "Kelly"),
                    Science = c(99, 100, 99, 100),
                    Music = c(19, 18, 20, 20),
                    Art = c(20, 20, 19, 18))
df2
```

Since `df1` and `df2` share the `Name` column, we can merge these two dataframes `by` that column: 

```{r, eval=FALSE}
?merge
```
```{r}
df_merge = merge(df1, df2, by = "Name")
df_merge
#What happened here? 
```

## Merging data - `cbind()` and `rbind()`

Other useful functions include `cbind()` and `rbind()`. `cbind()` will bind two data frames by their columns. Let's cbind `df1` and a new dataframe, `df3`:
```{r, eval=FALSE}
?cbind
?rbind # Click the "Combine R Objects by Rows or Columns" link
```
```{r}
df3 = data.frame(Name = c("Heather", "Billy", "Hector", "Jane"),
                  City = c("San Francisco", "Los Angeles", "San Francisco", "Chico"),
                  Math = c(49, 44),
                  Reading = c(10,10))
df3

df_cbind = cbind(df1, df3)
df_cbind
```

> NOTE: cbind will not warn you if you have duplicate column names! Which columns have been duplicated? 

`rbind()` will add more rows to an existing data frame. An example:

```{r}
dfB = data.frame(Name = c("James", "Tanisha", "Elizabeth", "Barack"),
                  City = c("Cleveland", "Memphis", "Detroit", "Chicago"),
                  Math = c(44, 49, 50, 49),
                  Reading = c(9, 9, 9, 10))

df_rbind = rbind(df1, dfB)
df_rbind
```
> NOTE: for rbind to work, you should have the same column names for both of your data frames! 

# **Challenge 6** 

1. Subsetting review. Use `gap` to create a subsetted dataframe named `UnitedStates_Mexico` that contains only data for countries USA and Mexico. (hint: does "and" actually mean "and" here?)  
2. Create a subsetted data frame that only contains only the country USA and lifeExp greater than 76! (hint: what is the meaning of "and" here in contrast to question #1?)  

```{r}
## YOUR CODE HERE
```

You can also order your data using `order`. For example, sort `mtcars` by ascending cyl and descending mpg: 
```{r, eval=FALSE}
data(mtcars)
new_cars = mtcars[order(mtcars$cyl, -mtcars$mpg),] 
new_cars
```
 
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>