---
title: "R-Fundamentals Part 1 solutions"
author: "Evan"
date: "March 25, 2019"
output: html_document
---

# Part 1 solutions

## Part 1 - Challenge 1
1. What is the three-piece recipe for variable definition? What piece goes where? Why?  
2. Define three numeric variables. Write a mathematical expression that uses only these variable names (not the actual numbers).  
3. Remove one of your numeric variables. Does your expression still work? Why or why not?  
4. Wipe you global environment clean.  

```{r}
# 1. 
# A) Variable name goes on left side of the equals sign
# B) Equals sign goes in the middle
# C) Code to be assigned goes on the right side of the equals sign

# 2. 
pi
radius = 2.34
height = 4.56

cylinder_volume = pi * (radius ** 2) * height
cylinder_volume

# 3. 
rm(radius)

# 4. 
rm(list = ls()) # Or click broom icon!

```

## Part 1 - Challenge 2
1. Define two numeric variables.  
2. Define two character variables.  
3. Define two logical variables.  
4. How do you check what types of data these variables are?  
5. What does the `ls()` function do? Where else do you see this information?
6. Can you perform math across these three data classes? Why or why not? 

```{r}
# 1. 
x = 5
y = 10

# 2. 
name = "Rex"
city = "Berkeley"

# 3. 
logical1 = TRUE
logical2 = FALSE

# 4. 
class(x)
class(y)
class(name)
class(city)
class(logical1)
class(logical2)

# 5
# ls() displays the variables saved in your global environment. Also see this information in your "Environment" tab in the upper right pane. 
ls()

# 6. 
# You can perform on the numeric variables, logical variables, or numeric/logical variables, but NOT the character data. 

x ** y
logical2 - logical1
x ** -logical1

x + name
```

## Part 1 - Challenge 3
1. Other `as.` ("as dot") functions exist as well, such as `as.numeric`, `as.character`, `as.logical`, and `as.factor`.  
2. Using your numeric, character, and logical variables from above:  
  2a. Can you convert numeric to integer type?  
  2b. Convert numeric to logical?  
  2c. Convert numeric to character?  
  2d. Convert logical to character?  
  2e. Convert character to numeric?  
  
```{r}
# 1. 

# 2
x; y; name; city; logical1; logical2

# 2a. 
numeric_to_integer = as.integer(x)
class(numeric_to_integer)
numeric_to_integer

# 2b. 
numeric_to_logical = as.logical(x)
class(numeric_to_logical)
numeric_to_logical # What happened here?

# 2c. 
numeric_to_character = as.character(x)
class(numeric_to_character)
numeric_to_character
  
# 2d. 
logical_to_character = as.character(logical1)
class(logical_to_character)  
logical_to_character

# 2e. 
character_to_numeric = as.numeric(name) # What happened here? 
```

## Part 1 - Challenge 4
1. Notice how `paste0()` is found within the `paste()` function - this is called "nesting functions". What is the purpose of the `paste0()` here? 

```{r}
# 1. 
# paste0() removes whitespace between characters. 
paste("North", "South", "East", "West")
paste0("North", "South", "East", "West")
```

## Part 1 - Challenge 5
1. `substr` can also be used to reassign parts of text. Fill in the question marks below so that that fox jumps over the cow instead of the dog. 

```{r}
# 1. 
pangram = paste("The", "quick", "brown", "fox", "jumps", "over", "the", "lazy", paste0("dog", "."))
pangram

substr(pangram, start = 41, stop = 43) = "cow"
pangram
```

## Part 1 - Challenge 6
1. Build a vector of logical data type and save it in a variable named `logical_vector`  
2. Can you divide `numeric_vector` by `numeric_vector2`? Why or why not?  
3. Can you multiply `numeric_vector` by `logical_vector`? Why or why not? 
4. Can you multiply `numeric_vector2` by `logical_vector`? 

```{r}
# 1. 
logical_vector = c(TRUE, FALSE, TRUE, FALSE, TRUE, TRUE, FALSE, FALSE)

numeric_vector = c(234, 31343, 78, 0.23, 0.0000002)
numeric_vector

numeric_vector2 = c(numeric_vector, c(0.1, 0.2, 0.3))
numeric_vector2

# 2. Yes, but...
numeric_vector / numeric_vector2

# 3. Yes, but...
numeric_vector * logical_vector

# 4. Yes! 
numeric_vector2 * logical_vector
```

What does the error message indicate for numbers 2 and 3?

> Warning message:
In numeric_vector/numeric_vector2 :
  longer object length is not a multiple of shorter object length

## Part 1 - Challenge 7
1. Save the "gdp" data frame as a .csv file named "gdp.csv". 

```{r}
write.csv(gdp, file = "gdp.csv")
```

## Part 1 - Challenge 8
```{r, eval=FALSE}
library(swirl)
swirl()
# follow the instructions until you can select number 1 "R Programming: The basics of programming in R"
```

Type `bye()` to exit swirl and return to RStudio. 
