---
title: "R-Fundamentals Part 3 solutions"
author: "D-Lab"
date: "3/30/2019"
output: html_document
---

# Part 3 solutions

## Part 3 - Challenge 1

1. Load the `iris` dataset.  
2. Run `summary`, `describe`, `describeBy`, and `table` on iris.  

```{r}
# 1. 
data(iris)

# 2. 
summary(iris)
summary(iris$Petal.Length)

describe(iris[, -5])

describeBy(iris[, -5], group = iris$Species)

table(iris$Species)
```

## Part 3 - Challenge 2

# **Challenge 1**
1. Create a histogram of one of the numeric variables from the `iris` dataset.  

```{r}
hist(iris$Petal.Length, 
     col = "goldenrod",
     main = "Histogram of Petal Length",
     xlab = "Peta Length (cm)",
     ylab = "Count", 
     xlim = c(1, 7),
     ylim = c(0, 40), 
     las = 1)
```

## Part 3 - Challenge 3

1. Create boxplots for one of the numeric variables in the `iris` dataset as parsed by Species.

```{r}
boxplot(iris$Petal.Length ~ iris$Species, 
        col = c("gray70", "gray50", "gray30"),
        main = "Iris petal lengths by species", 
        xlab = "Species",
        ylab = "cm",
        las = 1)
```

## Part 3 - Challenge 4

1. Using the `iris` dataset, plot two numeric variables in a scatterplot. 

```{r}
plot(x = iris$Petal.Length, 
     y = iris$Petal.Width, 
     main = "Scatterplot of iris petal lengths and widths",
     xlab = "Petal Length (cm)", 
     ylab = "Petal Width (cm)", 
     col = as.integer(iris$Species),
     pch = as.integer(iris$Species))
legend("topleft", inset = 0, title = "Species", cex = 1,
       c("setosa", "versicolor", "virginica"), 
       col = c(1,2,3), pch = c(1,2,3),
       horiz = FALSE)
```

## Part 3 - Challenge 5

1. Create a ggplot histogram of one of the variables from the sleep_VIM dataset. 

```{r}
sleep_VIM = read.csv("sleep_VIM.csv")

ggplot(sleep_VIM, aes(x = Gest)) + 
  geom_histogram(fill = "orange", bins = 20) + 
  ggtitle("Gestation lengths for mammals") + 
  xlab("Number of days") + 
  ylab("Frequency") + 
  theme_light()
```

## Part 3 - Challenge 6

1. Create ggplot boxplots for mpg for 4, 6, and 8 cylinder cars from the mtcars dataset.

```{r}
data(mtcars)
?mtcars
str(mtcars)

ggplot(mtcars, aes(x = as.factor(cyl), y = mpg, fill = as.factor(cyl))) + 
  geom_boxplot() + 
  ggtitle("Boxplots for mpg by cyl") + 
  xlab("Number of cylinders in engine") + 
  ylab("Miles per gallon") + 
  theme_dark() +  
  guides(fill = F)
```

## Part 3 - Challenge 7

1. In lines the ggplot code above, what are the arguments inside of our second "theme" argument doing?

```{r}
# Change legend position: top, bottom, left, right
theme(legend.position = "top",
      
      # Change horizontal justification and size of plot title
      plot.title = element_text(hjust = 0.5, size = 20),
      
      # Change size of the legend title
      legend.title = element_text(size = 10),
      
      # Change size of the legend text
      legend.text = element_text(size = 5),
      
      # Change the angle and horizontal justification of the x-axis text
      axis.text.x = element_text(angle = 45, hjust = 1)) 
```

## Part 3 - Challenge 8

1. What are the assumptions for a t-test?  
2. Fit a t-test for Petal Lengths between iris versicolor and virginica species.  
3. What does this result mean?  

```{r, eval = F}
?t.test

# 1. 
https://en.wikipedia.org/wiki/Student%27s_t-test#Assumptions

# 2. subset! 
versicolor = iris[iris$Species == "versicolor", ]
virginica = iris[iris$Species == "virginica", ]

t.test(versicolor$Petal.Length, virginica$Petal.Length)
```

## Part 3 - Challenge 9

1. What are the assumptions for a one-way analysis of variance?  
2. Use the `mtcars` dataset to perform an ANOVA and TukeyHSD test for mpg between the three cyl sizes (4, 6, and 8). 

```{r}
# 1. 
https://en.wikipedia.org/wiki/One-way_analysis_of_variance#Assumptions

# 2. 
data(mtcars)
mtcars_aov = aov(mpg ~ as.factor(cyl), data = mtcars)
summary(mtcars_aov)
TukeyHSD(mtcars_aov)
```
