---
title: "R-Fundamentals Part 4 solutions"
author: "D-Lab"
date: "3/30/2019"
output: html_document
---

# Part 4 solutions

# Example project

You are given a dataset you have never seen before. You want to use RStudio to import the dataset and learn about it a little before you proceed with preprocessing tasks. 

##### 1. Load the `mtcars` dataset.  
##### 2. What are some useful functions to learn more about the data? 
##### 3. Might you surmise a relationship about something like engine size and miles per gallon?  
##### 4. Create boxplots for mpg as parsed by cyl. 
##### 5. What does `cor.test` reveal about mpg and cyl? 
##### 6. Create a scatterplot of the "mpg" and "hp" variables using `ggplot()`. What happens?  
##### 7. Can cyl be used to predict mpg in a linear regression model? Why must you learn about the assumptions of a statistical test before applying it? 

```{r}
# 1. 
data()
data(mtcars)

# 2. 
?mtcars
str(mtcars)
head(mtcars)
mtcars
View(mtcars)

# field notes, lab reports, scientific instruments, Google, etc. 

# 3. 
# Larger engine equals less miles per gallon? 

# 4. 
boxplot(mpg ~ cyl, data = mtcars,
        col = c("red", "green", "gold"),
        main = "mtcars boxplots",
        xlab = "# of cylinders",
        ylab = "mpg", las = 1)

# or

plot_box = ggplot(mtcars, aes(cyl, mpg, 
                              fill = as.factor(cyl))) + 
  geom_boxplot() + 
  theme(legend.position = "top") + 
  labs(fill = "Cylinders")
plot_box

# 5. 
cor.test(mtcars$mpg, mtcars$cyl)

# 6. 
plot_scatter = ggplot(mtcars, aes(hp, mpg, color = mpg)) + 
  geom_point(size = 8, alpha = 0.8) +
  theme_minimal() 
plot_scatter

# 7. 
lm1 = lm(mpg ~ cyl, data = mtcars)
lm1
summary(lm1)
```
  
# Gapminder dataset
(in class example)

# Bonus 
### Challenge 1
1. Write a for loop that outputs something. 
```{r}
for (i in 1:10) {
  print(i / 12)
}

# or

for (time in c(12, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11)) {
  print(paste("The time is", time, "AM"))
}
```

### Challenge 2
```{r}
for (x in c(2,4,6,8,10,12)) {
  if (x < 8) {
    print(paste("Whoa the number is", x))
  }else{
    print("nope...")
  }
}

# or

for (i in 1:12) # what happened to the curly braces? 
    if (i < 10) {
        print(paste("The time is", i, "o'clock"))
        }else{
        print("It is getting late!")
}
```

### Challenge 3
Say we choose 25 people at random. What is the probability two or more of them have the same birthday?'
```{r}
## See the Part 4 .Rmd file for the answer
```
