---
title: "R-Fundamentals Part 2 solutions"
author: "D-Lab"
date: "3/30/2019"
output: html_document
---

# Part 2 solutions

## Part 2 - Challenge 1

1. Load the mammals "sleep_VIM" dataset from the data folder using `read.csv()`. Save it in a variable named "Sleep_VIM".  
2. Remember what `ls()` does? What does `dir()` do?  

```{r}
# 1. 
Sleep_VIM = read.csv("sleep_VIM.csv")

# 2. 

# ls() will show you the variables saved in your global environment
ls() 

# dir() will show you the files in your current working directory
dir()
```

## Part 2 - Challenge 2

1. Practice extracting a few variables from the `gap` dataframe using the dollar sign `$`  
2. Plot them in a histogram using the `hist()` function. How do you do this?  
3. Can all data types be visualized with histograms? Why or why not?  

```{r}
# 1. 
gap$country
gap$continent
gap$gdpPercap

# 2. 
hist(gap$country) # Error in hist.default(gap$country) : 'x' must be numeric
hist(gap$gdpPercap)

# 3. No, not all data types can be visualized with histograms. The error message above states that the input variable 'x' must be numeric! 
```

## Part 2 - Challenge 3

1. Load the `iris` dataset! Type `data(iris)` to load it.  
2. What is this dataset? How do you find out?  
3. Subset the `iris` dataset in three ways using column names and positive and negative integers.  
4. Subset `iris` using logical tests, `&`, and `|`    

```{r}
# 1. 
data(iris)

# 2.
?iris
str(iris)
https://en.wikipedia.org/wiki/Iris_flower_data_set

# 3. 
iris_varnames = iris[ , c("Sepal.Length", "Species")]
head(iris_varnames)

iris_posint = iris[ , c(1, 5)]
head(iris_posint)

iris_negint = iris[ , -c(1, 5)]
head(iris_negint)
```

## Part 2 - Challenge 4

Subset `gap`:  
1. Once in one dimension.  
2. Twice in two dimensions using either bracket notation `[ , ]` or `subset`.  
3. Thrice using logical tests.  

```{r}
# 1. 
gap$lifeExp

# 2. 
gap_bracket_notation = gap[c(1:3, 1704), 1:3]
gap_bracket_notation

gap_subset_function = subset(gap, 
                             subset = country == "Australia" | country == "New Zealand", 
                             select = c(1,2,4))
gap_subset_function

# 3. 
gap_logical1 = gap[gap$continent == "Oceania", ]
gap_logical1

gap_logical2 = gap[gap$country == "Mexico" & gap$year == 2007, ]
gap_logical2

gap_logical3 = gap[gap$country == "Mexico" | gap$year == 2007, ]
gap_logical3
```

## Part 2 - Challenge 5

1. Subsetting review. Use `gap` to create a subsetted dataframe named `UnitedStates_Mexico` that contains only data for countries USA and Mexico. (hint: does "and" actually mean "and" here?)  
2. Create a subsetted data frame that only contains only the country USA and lifeExp greater than 76. Save this in a variable named `US_lifeExp`. (hint: what is the meaning of "and" here in contrast to question #1?)  

```{r}
# 1. Here, "and" actually means we need to use the "or" | operator to select data for either United States or Mexico. 
UnitedStates_Mexico = gap[gap$country == "United States" | gap$country == "Mexico", ]
UnitedStates_Mexico

# 2. Here, we must use the "and" & operator so that both conditions (United States, lifeExp > 76) are satisfied. 
US_lifeExp = gap[gap$country == "United States" & gap$lifeExp > 76, ]
US_lifeExp
```
