---
title: "R-Fundamentals Part 1 solutions"
author: "Evan"
date: "March 25, 2019"
output: html_document
---

# Part 1 solutions

## **Challenge 1**
What are functions and arguments? Describe a few examples of each.  

* Functions perform actions on things!  
  * Examples: `mean()`, `plot()`, `lm()`, and `summary()`.   
* Arguments are the things - values, mathematical expressions, text, datasets, etc.  
  * Examples: type `?mean`, `?plot`, `?lm`, and `?summary` and look at the "arguments" section.  
  
## **Challenge 2**
1. What is the three-piece recipe for variable definition? What piece goes where? Why?  
2. Define two numeric variables. Write a mathematical expression that uses only these two variables (not the actual numbers).  
3. Remove one of these variables. Does your expression still work? Why or why not?  
4. Wipe you global environment clean.    

```{r}
# 1. 
# A) Variable name goes on left side of the equals sign
# B) Equals sign goes in the middle
# C) Code to be assigned goes on the right side of the equals sign

# 2. 
pi
radius = 2.34
height = 4.56

cylinder_volume = pi * (radius ** 2) * height
cylinder_volume

# 3. 
rm(radius)

# 4. 
rm(list = ls()) # Or click broom icon!

```

## **Challenge 3**
1. Define two numeric variables.  
2. Define two character variables.  
3. Define two logical variables.  
4. How do you check what class of data these variables are?  
5. What does the `ls()` function do? Where else do you see this information?
6. Can you perform math across these three data classes? Why or why not? 

```{r}
# 1. 
x = 5
y = 10

# 2. 
name = "Rex"
city = "Berkeley"

# 3. 
logical1 = TRUE
logical2 = FALSE

# 4. 
class(x)
class(y)
class(name)
class(city)
class(logical1)
class(logical2)

# 5
# ls() displays the variables saved in your global environment. Also see this information in your "Environment" tab in the upper right pane. 
ls()

# 6. 
# You can perform math on numeric/integer variables, and logical variables, but NOT the character variables. 

x ** y
logical2 - logical1
x ** -logical1
```
```{r eval = F}
x + name
```

## **Challenge 4**

1. Other `as.` ("as dot") functions exist as well, such as `as.numeric`, `as.character`, `as.logical`, and `as.factor`.  
2. Using your numeric, character, and logical variables from above:  
  2a. Can you convert numeric to integer type?  
  2b. Convert numeric to logical?  
  2c. Convert numeric to character?  
  2d. Convert logical to character?  
  2e. Convert character to numeric?  
  
```{r}
# 1. 
NA

# 2
x; y; name; city; logical1; logical2

# 2a. 
numeric_to_integer = as.integer(x)
class(numeric_to_integer)
numeric_to_integer

# 2b. 
numeric_to_logical = as.logical(x)
class(numeric_to_logical)
numeric_to_logical # What happened here?

# 2c. 
numeric_to_character = as.character(x)
class(numeric_to_character)
numeric_to_character
  
# 2d. 
logical_to_character = as.character(logical1)
class(logical_to_character)  
logical_to_character

# 2e. 
character_to_numeric = as.numeric(name) # What happened here? 
```

## **Challenge 5**
1. Notice that the `paste0()` function is found inside of `paste()`. What is the purpose of the `paste0` here? 

```{r}
# 1. 
pangram = paste("The", "quick", "brown", "fox", "jumps", "over", "the", "lazy", paste0("dog", "."))
pangram
# paste0() removes whitespace between characters. Change paste0() to simply paste() - what happened?
paste("N", "S", "E", "W") # cardinal directions
paste0("N", "S", "E", "W")
```

## **Challenge 6**
1. `substr` can also be used to reassign parts of text. Fill in the question marks below so that that fox jumps over the cow instead of the dog. 

```{r}
# 1. 
pangram = paste("The", "quick", "brown", "fox", "jumps", "over", "the", "lazy", paste0("dog", "."))
pangram

substr(pangram, start = 41, stop = 43) = "cow"
pangram
```

## **Challenge 7**
1. Build three vectors: 1) a logical vector named `logical_vector`, 2) a numeric vector named `numeric_vector`, and 3) another numeric vector named `numeric_vector2`. 
2. Can you divide `numeric_vector` by `numeric_vector2`? Why or why not?  
3. Can you multiply `numeric_vector` by `logical_vector`? Why or why not? 
4. Can you multiply `numeric_vector2` by `logical_vector`? Why or why not?

You might get a different answer than your neighbors. What is [Recycling of Vectors in R](http://eriqande.github.io/rep-res-web/lectures/vectorization_recycling_and_indexing.html#recycling)? 

```{r}
# 1. 
logical_vector = c(TRUE, FALSE, TRUE, FALSE, TRUE, TRUE, FALSE, FALSE)

numeric_vector = c(234, 31343, 78, 0.23, 0.0000002)
numeric_vector

numeric_vector2 = c(numeric_vector, c(0.1, 0.2, 0.3))
numeric_vector2

# 2. Yes, but...
numeric_vector / numeric_vector2

# 3. Yes, but...
numeric_vector * logical_vector

# 4. Yes! 
numeric_vector2 * logical_vector
```

What does the error message indicate for numbers 2 and 3?

> Warning message:
In numeric_vector/numeric_vector2 :
  longer object length is not a multiple of shorter object length

## **Challenge 8**
Where have you seen something that looks like a dataframe before?  

[Microsoft Excel](https://support.office.com/en-us/article/excel-for-windows-training-9bc05390-e94c-46af-a5b3-d7c22f6990bb)

[Download MS Excel](https://products.office.com/en-us/excel)  

[D-Lab Basics of Excel](https://github.com/dlab-berkeley/Basics-of-Excel)

[D-Lab Visualization in Excel](https://github.com/dlab-berkeley/visualization-in-Excel)

## **Challenge 9**
Read ahead to the Part2 materials to learn about subsetting lists and matrices! In 2-3 minutes of reading/skimming, how do you think subsetting works? 

1) dollar sign `$` operator,  
2) bracket `[ , ]` notation, and 
3) `subset()` function

... are just a few ways! 

## **Challenge 10**
1. Save the "gdp" data frame as a .csv file named "gdp.csv". 

```{r eval = F}
write.csv(gdp, file = "gdp.csv")
```

## **Challenge 11**
```{r, eval=FALSE}
library(swirl)
swirl()
# follow the instructions until you can select number 1 "R Programming: The basics of programming in R"
```

Type `bye()` to exit swirl and return to RStudio. 
