---
title: "R-Fundamentals solutions"
author: "Evan"
date: "March 25, 2019"
output: html_document
---

> Remember to click "Code" --> "Show document outline" to see a clickable list of the answers.  

# Part 1

## Part 1 - Challenge 1
1. What is the three-piece recipe for variable definition? What piece goes where? Why?  
2. Define five numeric variables. Write a mathematical expression that uses only these five numeric variable names (not the actual numbers).  
3. Remove one of your numeric variables. Does your expression still work? Why or why not?  
4. Wipe you global environment clean.  


## Part 1 - Challenge 2
1. Define two numeric variables.  
2. Define two character variables.  
3. Define two logical variables.  
4. How do you check what types of data these variables are?  
5. What does the `ls()` function do? Where else do you see this information?
6. Can you perform math across these three data classes? Why or why not? 


## Part 1 - Challenge 3
1. Other `as.` ("as dot") functions exist as well, such as `as.numeric`, `as.character`, `as.logical`, and `as.factor`.  
2. Using your numeric, character, and logical variables from above:  
  2a. Can you convert numeric to integer type?  
  2b. Convert numeric to logical?  
  2c. Convert numeric to character?  
  2d. Convert logical to character?  
  2e. Convert character to numeric?  
  
  
## Part 1 - Challenge 4
1. Notice how `paste0()` is found within the `paste()` function - this is called "nesting functions". What is the purpose of the `paste0` here? 


## Part 1 - Challenge 5
1. `substr` can also be used to reassign parts of text. Fill in the question marks below so that that fox jumps over the cow instead of the dog. 


## Part 1 - Challenge 6
1. Build a vector of logical data type and save it in a variable named `logical_vector`  
2. Can you divide `numeric_vector` by `numeric_vector2`? Why or why not?  
2. Can you multiply `numeric_vector` by `logical_vector`? Why or why not? 


## Part 1 - Challenge 7
1. Save the "gdp" data frame as a .csv file named "gdp.csv". 


## Part 1 - Challenge 8
```{r, eval=FALSE}
library(swirl)
swirl()
# follow the instructions until you can select number 1 "R Programming: The basics of programming in R"
```

Type `bye()` to exit swirl and return to RStudio. 


# Part 2

## Part 2 - Challenge 1
## Part 2 - Challenge 2
## Part 2 - Challenge 3
## Part 2 - Challenge 4
## Part 2 - Challenge 5
## Part 2 - Challenge 6
## Part 2 - Challenge 7
## Part 2 - Challenge 8

# Part 3

## Part 3 - Challenge 1
## Part 3 - Challenge 2
## Part 3 - Challenge 3
## Part 3 - Challenge 4
## Part 3 - Challenge 5
## Part 3 - Challenge 6
## Part 3 - Challenge 7
## Part 3 - Challenge 8

# Part 4
# Example project

You are given a dataset you have never seen before. You want to use RStudio to import the dataset and learn about it a little before you proceed with preprocessing tasks. 

##### 1. Load the `mtcars` dataset.  
##### 2. What are some useful functions to learn more about the data? 
##### 3. Might you surmise a relationship about something like engine size and miles per gallon?  
##### 4. Create boxplots for mpg as parsed by cyl. 
##### 5. What does `cor.test` reveal about mpg and cyl? 
##### 6. Create a scatterplot of the "mpg" and "hp" variables using `ggplot()`. What happens?  
##### 7. Can cyl be used to predict mpg in a linear regression model? Why must you learn about the assumptions of a statistical test before applying it? 

```{r}
# 1. 
data()
data(mtcars)

# 2. 
?mtcars
str(mtcars)
head(mtcars)
mtcars
View(mtcars)

# field notes, lab reports, scientific instruments, Google, etc. 

# 3. 
# Larger engine equals less miles per gallon? 

# 4. 
boxplot(mpg ~ cyl, data = mtcars,
        col = c("red", "green", "gold"),
        main = "mtcars boxplots",
        xlab = "# of cylinders",
        ylab = "mpg", las = 1)

# or

plot_box = ggplot(mtcars, aes(cyl, mpg, 
                              fill = as.factor(cyl))) + 
  geom_boxplot() + 
  theme(legend.position = "top") + 
  labs(fill = "Cylinders")
plot_box

# 5. 
cor.test(mtcars$mpg, mtcars$cyl)

# 6. 
plot_scatter = ggplot(mtcars, aes(hp, mpg, color = mpg)) + 
  geom_point(size = 8, alpha = 0.8) +
  theme_minimal() 
plot_scatter

# 7. 
lm1 = lm(mpg ~ cyl, data = mtcars)
lm1
summary(lm1)
```
  
# Gapminder dataset
(in class example)

# Bonus 
### Challenge 1
1. Write a for loop that outputs something. 
```{r}
for (i in 1:10) {
  print(i / 12)
}

# or

for (time in c(12, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11)) {
  print(paste("The time is", time, "AM"))
}
```

### Challenge 2
```{r}
for (x in c(2,4,6,8,10,12)) {
  if (x < 8) {
    print(paste("Whoa the number is", x))
  }else{
    print("nope...")
  }
}

# or

for (i in 1:12) # what happened to the curly braces? 
    if (i < 10) {
        print(paste("The time is", i, "o'clock"))
        }else{
        print("It is getting late!")
}
```

### Challenge 3
Say we choose 25 people at random. What is the probability two or more of them have the same birthday?'
```{r}
## See the Part 4 .Rmd file for the answer
```

