---
title: 'Part5-2: functions'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: 
  html_document: 
    number_sections: yes
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---

# Functions

In R, a function is a set of code that might be more useful if it is self-contained and/or is to be used repeatedly. For example, if you want to run the same satatistical tests on several datasets, you might want to write a function that contains instructions for all the tests once, so that you do not have to rewrite the code each time.  

Every function has four parts and its basic syntax looks like this:  

> function.name = function(x){
>   body of function
> }

1. `function_name` - like saving variables in R, it is helpful if you give your function a relevant name.  

2. `function(x)` - `function` lets R know you are writing a function and `(x)` contains states that there is one input/argument, x.  

3. `{body of function}` - is contained within curly braces { } and denotes the statements that you want the function to evaluate.  

4. The `environment` (global environment) that the function operates within.  

## Functions - Example 1

Write a simple `test_function` that squares a given number x: 

```{r}
test_function = function(x){
  x ^ 2
}

class(test_function)        # Returns the class of `test_function`
formals(test_function)      # Shows the defined arguments
body(test_function)         # Displays the statements to be evaluated
environment(test_function)  # Returns the "global" environment
test_function               # Shows your function as you have written it
```

See how our function has `R_GlobalEnv` as it's environment? That's because we defined it in the global environment. This means that if you tell a function to look for a variable, it will look in the global namespace - not within the function itself.

To evaluate this function, call it with your argument `x` within parantheses. R then evaluates the body of the function and returns the desired output for the specified input. 

If we want to see the result when `x` = 2, we type the value as the argument into our function:

```{r}
test_function
test_function(2)

# Also works on vectors and other data stuctures! 
test_function(1:10) # is this the same as c(1:10) ^ 2?
```

## Functions - Example 2 - default arguments

We can also add a predetermined value for one or more arguments, which serves as a default value that we can change in particular applications. You can change the value of "y" - it is an optional argument because it is named and it has a default value. By convention, required arguments do not have default values and are listed first in the function definition.

```{r}
f = function(x, y=2){
  (x + y) / y
}

f(x = 5)
f(x = 5, y = 10)
```

## Functions - Example 3 - function environment (scope)

Let's try one more example where we define variable `z` within the function itself: 

```{r}
f = function(x, y=2){
  z = 5
  (x + y) / z
}
# since we defined two arguments x and y, the function thinks the first one is x and the second one is y. You can overwrite y like this:
f(4, 4)  

f(x = 4, y = 4) # produces the same result

ls() # objects defined in the function environment only exist inside the function; they are not found in the global environment! 
```

Now we ask: does `z` appear in your global environment? Why not? `z` is protected from what is happening outside the function!  

Now define `z` in the global environment and see if the function can access it:

```{r}
z = 5 # define z outside of the function environment
f = function(x, y=2){
  (x + y) / z
}
f(4, 4)
ls() # the function envrionment can indeed use objects defined in the global environment. 
```
`z` is now defined in the global environment and the function can pull variables from this global namespace.  

For example, when using constants that never change, you might want to consider defining them in the global namespace. If it is unique to the function, keep it in the function! Scope is tricky!  

## Functions - Example 4 - for-loop inside of function
```{r}
function_squared = function(x){
  for (i in 1:x) {
    y = i ^ 2
  print(y)
  }
}
function_squared(10)
```

## Functions - Example 5 - using a data frame

You can even use the columns of a data frame as your inputs: 

```{r}
# construct a dummy data frame 
cylinders = data.frame(height = c(3,4,5,6,7),
                       radius = c(1,3,2,1,5))
cylinders

# write a function that computes the volume of these cylinders
# the two arguments are defined as the height and radius columns
cylinders_function = function(height, radius){
  volume = pi * radius ^ 2 * height
  print(volume)
}

str(cylinders)

# plug in height and radius
cylinders_function(height = cylinders$height, radius = cylinders$radius)
```

# **Challenge 3**

Describe what is happening in the below code for a simple combination lock test: 

```{r}
# combination lock example

# 1.  
set.seed(123)
combos = paste0(sample(1:60,100, replace = T), "-",  # paste0 or paste?
              sample(1:60,100, replace = T), "-", 
              sample(1:60,100, replace = T))
combos

# 2. 
lock = function(x) {
for(x in 1:length(combos)){
  if(combos[x] == "13-24-48"){
    print(paste(combos[x], ":", "unlock"))
}else{
  print(paste("ERROR", ":", combos[x]))
}
}
}

# 3. 
lock(combos)
```
