---
title: 'R Fundamentals Part3-4: ggplot2'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# ggplot2

The "ggplot2" R package is an efficient way to visualize data and its syntax is easier to navigate than the base plotting functions. The [ggplot2](http://www.sthda.com/english/wiki/ggplot2-essentials) package is part of the [tidyverse](https://www.tidyverse.org/) - a series of R packages designed to simplify R syntax and functionality. "gg" stands for ["Grammar of Graphics."](https://en.wikipedia.org/wiki/Leland_Wilkinson).  

Although the package is called "ggplot2", `ggplot` is the main function call. You need three things to create a ggplot:  

1) a dataset - some data to be visualized.  

2) "aes"thetics - definitions of your coordinate system, point color and shape mappings, etc.  

3) geoms - how your data should be represented: points, bars, densities, ribbons, etc.  

`ggplot` works in layers. Each time you wish to add a geom, title, axis label, or background color, you must add a new layer by typing the plus `+` sign at the end of the previous line.   

Check out the help files: 

```{r}
library(ggplot2)
?ggplot2
data(iris)
```

# ggplot histogram 

Remember how we said the `ggplot` function works in layers? Look what happens if we specifiy our data set ("iris") and our global aesthetic ("aes"), but we do not specify our "geom". We get the coordinate system and a defined X axis (lefExpt), but no representation of the data!

```{r}
ggplot(data = gap, aes(x = lifeExp))
```

We can add the bars by adding a new layer with `+`, and then specify "geom_histogram()" to produce the histogram! 

To see the list of available geoms, type "geoms_" and press the `TAB` key. 

```{r}
ggplot(data = gap, aes(x = lifeExp)) + 
    geom_histogram()
```

You can also change the color of the bars by adding the `fill` argument to the global aes, change the background color by adding `theme_` in a new layer, and the title by adding `ggtitle` in a new layer. The `bins` argument will change the number and width of the histogram bars.

```{r}
ggplot(data = gap, aes(x = lifeExp)) + 
  geom_histogram(fill = "blue", color = "black", bins = 10) + 
  ggtitle("Life expectancy for the gap dataset") + 
  xlab("Life expectancy (years)") + 
  ylab("Frequency") + 
  theme_classic() 
```

# **Challenge 5**
1. Create a ggplot histogram of one of the variables from the sleep_VIM dataset. 

```{r}
## YOUR CODE HERE
```

# ggplot boxplots

Boxplots continue to follow this logic. However, this time we should specifically define our X axis as the factor variable we wish to investigate, and the Y axis as some numeric value: 

```{r}
ggplot(data = gap, aes(x = continent, y = lifeExp, fill = continent)) + 
  geom_boxplot() + 
  ggtitle("Boxplots for lifeExp by continent") + 
  xlab("Continent") + 
  ylab("Life expectancy (years)") +
  theme_minimal() # + 
  # guides(fill = FALSE) 
```

What happens if you un-hashtage `guides(fill = FALSE)` and the plus sign in lines 68 and 69 above? 

# **Challenge 6**

1. Create ggplot boxplots for 4, 6, and 8 cylinder cars from the mtcars dataset.

```{r}
## YOUR CODE HERE
```

# ggplot scatterplots

Scatterplots work similarly. However, this time we must define both our X and Y axes in our global "aes" and select "geom_point()" as our geom. We can then automatically map the factor variable ("Species") to point types and shapes in the `color` and `shape` arguments. By doing so, a legend is automatically created for us! This is a strong advantage over base scatterplots.  

The other layer arguments apply as well. `size` will change the size of the points, while `alpha` will change their transparency. 

We can also include a second `theme` argument layer where can can specify the legend position via `legend.position`. 

The X and Y axis labels can be changed via `xlab` and `ylab` layers, for example if we want to include our units of measurement (centimeters):

```{r}
ggplot(data = gap, aes(x = lifeExp, y = gdpPercap, color = continent, shape = continent)) + 
    geom_point(size = 5, alpha = 0.5) + 
    theme_classic() +
    ggtitle("Scatterplot of life expectancy by gdpPercap") +
    theme(legend.position = "top") + 
    xlab("Life expectancy (years)") + 
    ylab("gdpPercap (USD)") + 
    theme(legend.position = "top",
          plot.title = element_text(hjust = 0.5, size = 20),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 5),
          axis.text.x = element_text(angle = 45, hjust = 1)) 
```

# **Challenge 7** 

1. In lines the ggplot code above, what are the arguments inside of our second "theme" argument doing?

```{r}
## 
```

There are many more additional arguments that you can use to tweak your ggplots. [Check out this cheatsheet to learn more!](https://www.rstudio.com/resources/cheatsheets/#ggplot2)