---
title: 'R Fundamentals Part3-2: data summarization'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# Data summarization

This section will cover a few ways to quickly produce descriptive statistics.  

## `summary`

`summary` provides six-number summaries for numeric/integer variables, and frequencies for factor/logical variables:

```{r}
gap = read.csv("gapminder-FiveYearData.csv")

summary(gap)

# or of a single vector
summary(gap$lifeExp)
summary(gap$continent)
```

# `describe` and `describeBy`

## `describe`

`describe` and `describeBy` from the `psych` R package provide some other metrics. We are going to subset `gap` so that it only includes the numeric variables within the `describe` call. 

Remember, we installed the 'psych' package on Day 1, so all we have to do is call it into our environment with `library`:

```{r}
library(psych)

# This works, but we see some asterisks next to our factor/categorical variables. 
psych::describe(gap) 

# Let's remove country, year, and continent - looks a little nicer!
describe(gap[ , -c(1,2,4)]) 
```

We can also subset these outputs to include only the measurements we want. First, save our describe object in a variable: 
```{r}
gap_describe = describe(gap[ , -c(1,2,4)]) 
gap_describe
```

Now we can subset it like we would a regular data frame. If we just want the values for "mean", "median", "sd", "skew", and "kurtosis" we could subset our data frame just like you learned in Part 2:

```{r}
gap_simple <- gap_describe[ , c("mean", "median", "sd", "skew", "kurtosis")]
gap_simple
```

This is convenient to quickly write output to file and copy/paste into your paper or presentation: 

```{r, eval=FALSE}
write.csv(gap_simple, "gap_simple.csv", row.names = TRUE)
#`row.names=TRUE` ensures that row names "pop", "lifeExp", and "gdpPercap" are included! 
```

Check your working directory for the new .CSV file! 
```{r}
dir()
```

## `describeBy`

We can also describe our data by a grouping variable using `describeBy()`. What if we want summary statistics not for the whole sample like above, but for each continent instead? 

```{r, eval=FALSE}
# Output summary statistics by one grouping variable:
summary_sub <- describeBy(gap[ , -c(1,2,4)], group = gap$continent)
summary_sub

# If we just want to view Africa, Asia, or Oceania, we can type:
summary_sub$Africa
summary_sub$Asia
summary_sub$Oceania

# We can also view just the means for Asia, we can type:
summary_sub$Asia[["mean"]] # or
summary_sub$Asia[[3]]

# If we just want the second value (lifeExp mean) for Asia (60.0649) we can type:
summary_sub$Asia[["mean"]][2]

# Or the medians for Oceania:
summary_sub$Oceania[[5]]
summary_sub$Oceania[["median"]]
summary_sub$Oceania[["median"]][1] # just the first value (pop)
```

## `table`

We can view frequencies for of categorical data like `gap$continent` with `table`

```{r}
# Return frequencies for the different continents
table(gap$continent) 

# Get frequencies for the different continents by year
table(gap$continent, gap$year) 
# Or 
table(gap$year, gap$continent) 
```

# **Challenge 1**

1. Load the `iris` dataset.  
2. Run `summary`, `describe`, `describeBy`, and `table` on iris.  

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>