---
title: 'R Fundamentals Part3-1: Part2 review and Part3 objectives'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output:
  html_document:
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 2 review

1. Set your working directory to the R-Fundamentals folder

```{r}
getwd()
## "Session" --> "Set working directory" --> "Choose directory"
## Choose the data-raw folder
getwd()
```

2. Load the `gap` data frame

```{r}
gap = read.csv("gapminder-FiveYearData.csv")

str(gap)

head(gap)
```

3. What is one dimensional subsetting? 

```{r}
## Examples go here
```

4. What is two dimensional subsetting? 

```{r}
## Bracket notation examples

## subset() examples
```

Part 3 Learning objectives  
1. An example workflow  
2. Data exploration - summarize and visualize  
3. Visualizations with ggplot2 
4. What is hypothesis testing?  
5. Fast intro to statistical testing  

# Data exploration and analysis in R

Research design, data collection, exploration, visualization, and analysis are fundamental aspects of research. Learning how to explore and analyze data will help you "think with the data" so that you become able to formulate better research designs and more efficient data collection protocols for your own research.  

Furthermore, when you read peer-reviewed work in your field, having an understanding of basic summaries, plots, and statistical tests will help you better grasp arguments that the authors are making and will allow you to more critically evaluate their rationale. This can be used to contextualize tests on your own data.  

Think about this process in terms of a [workflow](https://en.wikipedia.org/wiki/Scientific_workflow_system), but we will just start with the research aspects today. It is common to have some _a priori_ defined research questions, hypotheses, and expectations of the data. You have learned a little R and have identified a problem to solve. It's always good to first define your purpose, goals, and aims clearly. The process might look like this:

# An example workflow

You will find that approaching research of all kinds generally falls into some standard buckets:  

1. Read the literature in your field. What suggested directions should you proceed in? Where are the gaps in the scientific knowledge? What is missing, is lacking, or otherwise needs to be expanded upon?  
2. Ask a well-informed research question using your literature review.  
3. Turn your research question into a set of testable hypotheses.  
4. Design data collection protocols/obtain data somehow.  
5. Import this data into R and clean/subset it.  
6. Compute descriptive statistics.  
7. Generate visualizations.  
8. Your data summaries and visualizations should (hopefully) indicate what you should specifically test. Descriptive statistics and plots help check for any obvious (or not so obvious) signals or relationships. Statistical testing can then be employed to test such relationships.  

> DISCLAIMER: Remember that we are not formally teaching you statistics, but instead the basics of responsible statistical programming application in R.  

# **Challenge 1** 

Approximately where are you in your own research workflow? 

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>