---
title: "R Fundamentals Part1-3: data classes and coercion"
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# Variable class

Everything in R has a class (or "type"). This is important because it determines what we can do to that thing! **The five most common data classes are: numeric, integer, character, logical and factor.** Use the `class` function to check a variable's type. 

## Numeric data

```{r}
number = 5
class(number)
```

Numeric type is the default class for all numbers in R, even those that look like integers! "numeric" type in R is synonymous with "float" or "double".  

## Character data

Let's try another example, this time using character data. Note that character data are **always** contained within quotation marks `" "`. Single or double quotes are fine as long as you are consistent. 

```{r}
welcome = "Welcome to the D-Lab"
ls()
class(welcome)

# Single and double quotes work similarly
contraction = 'I am hungry.'
contraction = "I am hungry."

# You can nest single quotes inside of double quotes
contraction = "I'm hungry"

# Or, you can use all single quotes along with escape characters
contraction = 'I\'m hungry'
```

"character" type is synonymous with "string" or "text" data. 

## Logical data

Logical data consists of `TRUE` or `FALSE` (or abbreviated as `T` and `F`. These are stored internally 1 (`TRUE`) and 0 (`FALSE`) and as such take on mathematical properties. 

```{r}
logical_variable = TRUE
logical_variable
class(logical_variable)

logical_variable + 2
logical_variable - 4
```

Logical values can also be used to test certain conditions useful for tasks such as subsetting data. You will learn more about these in Part 2.  

```{r}
FALSE < TRUE # less than
TRUE >= TRUE # greater than or equal to
TRUE == TRUE # equivalent to (equal to)
TRUE == FALSE
FALSE == FALSE
FALSE != FALSE # not equivalent to (not equal to)
FALSE != TRUE

TRUE & TRUE # boolean "and"
TRUE & FALSE

TRUE | TRUE # boolean "or"
TRUE | FALSE
```

You will also find that logical values are also used to turn certain arguments on or off: 

```{r}
num_vec = c(2, 5, 8, NA, NA)
num_vec

mean(num_vec, na.rm = FALSE) # Does not work. 

mean(num_vec, na.rm = TRUE) # Works properly. 

(2 + 5 + 8) / 3 # Do it manually! (sanity check)
```

`c` is the function to contacenate (combine) a vector of data (see "Vector" section below).  

`NA` is how R recognizes missing data (see Part 2 materials).  

The `mean` function computes the arithmetic mean of a vector of numbers. 

The `na.rm = ` argument is first set to `FALSE` and does not remove the missing values before computation proceeds. We can define the `na.rm = ` argument to `TRUE` so that missing values are removed prior to computation and the correct output is produced.  

# **Challenge 3**

1. Define two numeric variables.  
2. Define two character variables.  
3. Define two logical variables.  
4. How do you check what class of data these variables are?  
5. What does the `ls()` function do? Where else do you see this information?
6. Can you perform math across these three data classes? Why or why not? 

```{r}
## YOUR CODE HERE
```

## Integer type and "type coercion"

Remember how even whole numbers default to numeric/float/double type? Turning one data type into another requires what is called [type coercion](https://en.wikipedia.org/wiki/Type_conversion), or changing one data class to another. 

Thankfully, R has functions to perform type coercion. What happens in the code below?

```{r}
num = 23
class(num)

int = as.integer(num)
class(int)
```

# **Challenge 4**

1. Other `as.` ("as dot") functions exist as well, such as `as.numeric`, `as.character`, `as.logical`, and `as.factor`.  
2. Using your numeric, character, and logical variables from above:  
  2a. Can you convert numeric to integer type?  
  2b. Convert numeric to logical?  
  2c. Convert numeric to character?  
  2d. Convert logical to character?  
  2e. Convert character to numeric?  

```{r}
## YOUR CODE HERE
```

# Factor type

Factor type is simply how R refers to [categorical data](https://en.wikipedia.org/wiki/Categorical_variable). You will learn more about factor types in later sections.  

# Vignettes

Double question marks ?? will lead you to coding walkthroughs called "vignettes". These usually come with preloaded data and examples:  

```{r}
??psych
```

Click "PDF" next to "psych::intro" to download the walkthrough. However, you will often find that your questions are not answered by neither the help pages nor vignettes. In that case you should Google-search your question or error message along with the name of a free help website. For example, to get help making a simple histogram, I would Google search "R histogram stack overflow". Take some time to browse the sites listed in the "Resources" section of the README.md file. 

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>