---
title: 'R Fundamentals Part1-2: variable assignment'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# Variable assignment

Variables are how we save things in R's memory so that they can be "called" (retrieved) when we need them. Assigning/defining these variables is called [variable assignment](https://en.wikipedia.org/wiki/Assignment_(computer_science)) and they are stored in your **global environment.** Traditionally, you can define objects using the **assignment operator** `<-`. This is a "less than" `<` symbol immediately followed by a hyphen `-`. You will see this symbol in many tutorials. 

> It's okay to use the `=` in place of the assignment operator. This works the same but uses less keystrokes.

> However, it's also important to note that `<-` and `= do not work in the same way. Consider [the following example](https://csgillespie.wordpress.com/2010/11/16/assignment-operators-in-r-vs/). You can see that `=` and `<-` have different scopes. 

```{r}
mean(x = 1:10) # You can't call x 
rm(x)
mean(x <- 1:10) # You can call x 
rm(x)
```

Also, the [Google R style guide](https://google.github.io/styleguide/Rguide.html) and Hadley Wickam's [style guide](http://adv-r.had.co.nz/Style.html) recommend `<-` for assignment. 

## Variable assignment requires three pieces of information  

1) object_name  
2) `=`  
3) code to be assigned  

Try it! Define a variable named `number` and assign to it the number 5:

```{r}
number = 5
```

"Call" (retrieve) the data contained within the variable by typing its name into your script and running the line of code. 

```{r}
number # 5 is returned
```

Note that `number` appears in your global environment tab! You can also use the `ls()` (list) function to see the values saved to your global environment.  

```{r}
ls()
```

Also use "tab completion" so that you only have to type the first few letters of your variable and press your tab key to complete it. 

```{r eval = F}
numb
```

Think of R as the fanciest of calculators. You can do all the maths with ease in R:

```{r}
number2 = 2
number3 = 23948723
number4 = 1239873498

number * number2 / number3 ** .5

# or 

number %*% number2 # matrix multiplication
number %% number2 # modulus division
number %/% number2 # integer division
```

## Naming R variables  

Variable names can be anything, but are always case sensitive. However, they cannot begin with a number and generally do not begin with symbols. Be consistent and give your variables brief descriptions of their contents. Names must be **unique.**  

If you reuse an old name, the old definition will be overwritten.  

Use an underscore as a placeholder for a space to create variable names with multiple words. Let's define and overwrite a variable to see how its definition changes.  

```{r}
welcome = "UC Berkeley" 
welcome

welcome = "Barrows Hall"
welcome
```

If we want to preserve the `welcome` variable, we should define a new object like so below: 

```{r}
welcome2 = "UC Davis"
welcome2
class(welcome2)
ls()
```

We now have multiple variables defined in our global environment! Note that `welome2` did not overwrite the variable `welcome`. 

Does the below naming conventions make sense? Remember, it is good to name your variables as brief descriptions of 

```{r}
this_is_a_pizza = 12
this_is_a_pizza

my_name = "Douglas Engelbart"

# learn more about him here: https://en.wikipedia.org/wiki/Douglas_Engelbart
```

## Removing variables

We remove individual variables from our environment with `rm()`. For example, to remove `welcome`, we type:

```{r eval = F}
rm(welcome)
ls()
welcome
```

See how `welcome` disappeared from our global environment? What happens if you try to call the `welcome` variable? What does the error message say? 

We can also wipe the entire environment with `rm(list = ls())` (or, click the broom icon in the upper right "Global Environment" pane)

```{r}
rm(list = ls()) 
ls()

```

All variables have been removed from our global environment.

# **Challenge 2**

1. What is the three-piece recipe for variable definition? What piece goes where? Why?  
2. Define two numeric variables. Write a mathematical expression that uses only these two variables (not the actual numbers).  
3. Remove one of these variables. Does your expression still work? Why or why not?  
4. Wipe you global environment clean.  

```{r}
## YOUR CODE HERE
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>