---
title: 'R Fundamentals Part1-4: data structures'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# Data structures

If assigning a single value to a variable is good, assigning **more than** one piece of data to a variable is better. R has many different functions to store more complex data structures as variables:  

1. `paste()` - used to concatenate (combine) text  
2. `c()` - contatenate vectors  
3. `list()` - create lists  
4. `matrix()` - construct matrices  
5. `data.frame()` - combine vectors into a sructured, tabular spreadsheet with rows and columns.  

# `paste` - combine words! 

R is very powerful for manipulating text data. Here are some basic functions for working with text. Use `paste()` to concatenate/combine character data. This will paste together separate words to larger texts.

```{r, eval=FALSE}
?paste
```
```{r}
pangram = paste("The", "quick", "brown", "fox", "jumps", "over", "the", "lazy", paste0("dog", "."))
pangram
```

Note here that R is a one-indexed language - output begins at the number 1 and is designated by a number 1 in square brackets on the left side of the console. The one and only element in this object is the sentence "The quick brown fox jumps over the lazy dog." Learn more about vectors below. 

# **Challenge 5**

1. Notice that the `paste0()` function is found inside of `paste()`. What is the purpose of the `paste0` here? 

```{r}
## YOUR CODE HERE
```

# Text manipulations: Function arguments in action

R also has some useful functions for character data such as `grep`, `substr`, `strsplit`, and `gsub`. Use your help function `?` to learn more! Recall that arguments go inside of the parentheses of R functions. Unnamed arguments are things like the number 4.  

`split = " "` is what is called a named argument. This argument goes inside the parentheses of another function such as `strsplit`. Refer back to our `pangram` variable:

## `strsplit` - splitting strings

```{r, eval=FALSE}
?strsplit
```
```{r}
words = strsplit(pangram, split = " ")
words 
```

A list of the nine individual words split by their blank spaces is returned. Learn more about how lists work below!

However, anything can be used as the separator. Define some slash separated text and split on the slash:

```{r}
slash_text = "This/is/slash/separated/text"
slash_text

slash_split = strsplit(slash_text, split = "/")
slash_split
```

Most functions require one or two arguments to be defined in order for it to properly run (not `ls()`!). You will find that functions are full of default, positional, and optional arguments that you can manipulate.  

## `substr` - subsetting strings

`substr` lets you extract text from certain character positions in character data. Refer back to `pangram` - extract just the 4th through the 11th characters.

```{r, eval=FALSE}
?substr
```
```{r}
sub_text = substr(pangram, start = 4, stop = 11)
sub_text
```

" quick b" is returned. 

# **Challenge 6**

1. `substr` can also be used to reassign parts of text. Fill in the question marks below so that that fox jumps over the cow instead of the dog. 

```{r eval = F}
## YOUR CODE HERE
substr(pangram, start = ?, stop = ?) = ?
```

How did you do? Counting out all the character positions (indexes or indices) is a pain! Thanksfully there are functions we can use to search for words. Note that our pangram is no longer a pangram!

## `gsub` - string substitution

You can also substitute with `gsub()`. Let's say we want to substitute "hippopotamus" for "cow" in `pangram`. We could type:

```{r eval = F}
pangram
hippo = gsub("cow", "hippopotamus", pangram)
hippo
```

## `grep` - pattern matching and replacement

```{r, eval=FALSE}
?grep
```
Learn more about the `grep` function in the data frame section below! Also see the resources section in the README.md file to learn about regular expressions. 

# Vector

A **VECTOR** is an ordered group of the same kind of data. "Ordered" means that position matters. Vectors are one-dimensional and homogenous, and are thus referred to by their type (e.g., character vector, numeric vector, logical vector). It does not matter what type of data are in the vector, as long as it is all the same! 

Create a numeric vector by combining/concatenating elements with `c()`  

```{r, eval=FALSE}
?c
```
```{r}
numeric_vector = c(234, 31343, 78, 0.23, 0.0000002)
numeric_vector
class(numeric_vector)
length(numeric_vector)

# turn off scientific notation
options(scipen = 999)
numeric_vector
```

This numeric vector contains five elements (it has a length of five).  

## Indexing vectors

You can refer to the position of a value within a data structure by calling its "index". Use square brackets to do so (more on this in the subsetting section of Part 2). To see what value occupies the third position of the object `numeric_vector`, you could type: 

```{r}
numeric_vector[3]
class(numeric_vector[3])
```

## Add items to vectors

You can also add items to a vector by adding them as additional arguments and separated by commas (as long as they are of the same type!)
```{r}
numeric_vector2 = c(numeric_vector, c(0.1, 0.2, 0.3))
numeric_vector2
```

# **Challenge 7**

1. Build three vectors: 1) a logical vector named `logical_vector`, 2) a numeric vector named `numeric_vector`, and 3) another numeric vector named `numeric_vector2`. 
2. Can you divide `numeric_vector` by `numeric_vector2`? Why or why not?  
3. Can you multiply `numeric_vector` by `logical_vector`? Why or why not? 
4. Can you multiply `numeric_vector2` by `logical_vector`? Why or why not?

You might get a different answer than your neighbors. What is [Recycling of Vectors in R](http://eriqande.github.io/rep-res-web/lectures/vectorization_recycling_and_indexing.html#recycling)? 

## `seq` generating (pseudo)random numbers  

You might need to create vectors that are sequences of numbers. You can do this via `seq`. Define the from, to, and by arguments: 

```{r, eval=FALSE}
sequence = seq(from = 0, to = 8, by = 1)
sequence
```

# Sequence where `by = 1` (`:`)

R also gives you a shorthand operator for creating sequences in whole number increments of 1. This is the colon symbol `:`

```{r}
0:8

sequence_object = c(28:36)
sequence_object 

0:length(numeric_vector)
```

## `set.seed`

You can also generate pseudo-random groups of numbers. Use `set.seed()` to ensure that we get the same pseudo-random draws from the parent universe, even on different computers or at different times!  

Let's generate five different vectors of length 10 for numeric, integer, logical, character, and factor data types. 

```{r eval = F}
?set.seed

# Sample numbers from the uniform distribution
?runif

# Sample numbers from the normal distribution. 
?rnorm

# Sample numbers, logical values, character strings, etc.
?sample
```

## (Pseudo) random sampling

Set the seed, and then choose our values:

```{r}
set.seed(1)

# 10 random samples from the uniform distribution between the numbers one hundred and one million
uniform = runif(n = 10, min = 100, max = 1000000) 
uniform
class(uniform)

# 10 random samples from the standard normal distribution with a mean of 0 and standard deviation of 1
normal = rnorm(n = 10, mean = 0, sd = 1) 
normal
class(normal)

# 10 random samples from between the numbers 1990 and 2010. 
# Note that `replace = TRUE` signifies that it is OK to reuse numbers already selected.
integer = sample(x = 1990:2010, size = 10, replace = FALSE) 
integer
class(integer)

# 10 random samples from a character vector
character = sample(c("Canada", "USA", "Mexico"), 10, replace = TRUE) 
character
class(character)

# 10 random samples from a logical vector
logical = sample(c(TRUE, FALSE), 10, replace = TRUE) 
logical
class(logical)
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>