---
title: 'R Fundamentals Part1-7: exporting and practice'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# Saving your work (`write.csv`)  

It is always good practice to organize, document, and save your work. Saving a dataframe as a .CSV file is a convenient way to store it for future use. **Anything saved will be placed into your working directory by default!** 

See where your data frame is saved by typing: 

```{r}
getwd()
```

> NOTE: we will begin Part 2 with a discussion about how to check and set the [working directory](https://en.wikipedia.org/wiki/Working_directory). The easiest way is to click "Session" --> "Set working directory" --> "Choose directory". Do this if you want to change your working directory location. Run `getwd()` again - did the file path change?  

The syntax looks like this:
`write.csv(object_name, file = "nameOfFile.csv", row.names = TRUE)`

# **Challenge 10**

1. Save the "gdp" data frame as a .csv file named "gdp.csv". 

```{r}
## YOUR CODE HERE
```

Check your working directory to see if it worked!  

> We will talk about saving plots as .PDF and .TIFF file types in Part 3!  

# **Challenge 11** Homework (optional)

swirl is a package that helps you learn R by using R. Visit the [swirl](http://swirlstats.com/) homepage to learn more. 

```{r, eval=FALSE}
library(swirl)
swirl()
# follow the instructions until you can select number 1 "R Programming: The basics of programming in R"
```

> NOTE: type `bye()` to exit swirl and return to RStudio. 

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>