---
title: 'R Fundamentals Part1-1: introduction'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output:
  pdf_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workshop goals

Part 1) Introduction to R, data types, data structures/data frames  
Part 2) Import, subset, and merge data; identifying missing data  
Part 3) Research design, data summarization and visualization  
Part 4) Statistical testing  

**Bonus**  
Part 5) For-loops and functions; Monte Carlo simulation, the "birthday problem"  
Part 6) What is RMarkdown?  
Part 7) Mini-project  

**Part 1 learning objectives:**  
1. Introduction: What is R? Navigating RStudio, variable assignment, global environment, packages, variable assignment, help  
2. Data types: numeric, integer, character, logical, and factor  
3. Data type coercion  
4. Data structures: vector, list, matrix, data frame  
5. Saving your work  
6. Swirl  

# Part 1 - Introduction

## The student

This document is designed as a resource for **the student**. Use these documents if you choose to review or read ahead.  

## The instructor

**The instructor** should emphasize that:  
1. R is free and open-source  
2. Is great for data visualizaiton  
3. Utilizes many analytical packages written for your field  
4. Is ideal for open and reproducible research  
5. Has strong community support and help and is used by virtually all academic disciplines, business, and industry  

Then, the instructor will:  
6. Provide motivating examples for the class  
7. Describe how to apply R to research  
8. Point out some of R's shortcomings along the way  
9. Lead typing walkthoughs  

# R and RStudio

R is a programming language based on the [S](https://en.wikipedia.org/wiki/S_(programming_language)) language and is [functional](https://en.wikipedia.org/wiki/Functional_programming), [object oriented](https://en.wikipedia.org/wiki/Object-oriented_programming), and [interpreted](https://en.wikipedia.org/wiki/Interpreted_language). These themes make R highly extensible. However, we do not need to think of R in such complicated terms. In this workshop, we provide you with basic programming competencies so you can immediately apply what you learn to your own research. In general, programming languages such as R can simply be thought of as a way to give instructions to a computer to accomplish a data task. R is the actual programming language you will learn to use.  

RStudio is the premier R graphical user interface (GUI) and integrated development environment (IDE) that makes R easier to use. It can directly interface with [GitHub](https://en.wikipedia.org/wiki/GitHub) for sharing, collaboration, and version tracking and even makes it easy to program in other languages such as [Bash](https://en.wikipedia.org/wiki/Bash_(Unix_shell)) and [Python](https://en.wikipedia.org/wiki/Python_(programming_language)). 

R's many additional user-defined software "packages" (additional software libraries that provide to you tricks and shortcuts) allow for relatively straightforward completion importing and preprocessing data, statistical testing, and producing publication-ready visualizations for statistical modeling, machine learning, deep learning, natural language processing, and time-series analysis.  

## Tools --> Global Options

Before we begin, let's set a few RStudio settings to improve your experience.  

1. Click "Tools --> Global Options --> Appearance" to change your color and font settings.  

2. Click "Tools --> Global Options --> Code" and check the box that says "Soft-wrap R source files" to wrap the text in your script to the width of the script pane.  

3. Click "Tools --> Global Options --> Code --> Display" and check the boxes that say "Highlight selected line" and "Highlight R function calls". 

# The basics: Navigating RStudio

Open RStudio! Then, open a new script by clicking "File --> New File --> R Script" or by pressing `Ctrl + Shift + N` (PC) or `command + shift + N` (Mac). After opening a new script, you should see four window "panes".  

## Top left pane (input/script)

Enter code in this savable `script` file in the top left pane. This is a plain text file but with a .R extension. Enter `2 + 2` in your script and run a line of code by pressing `command + enter` (Mac) or `Ctrl + enter` (PC). Or, click the "Run" button at the top of the script.  

A hashtag `#` tells R that you do not want that particular line or block of code to be run - this is called **commenting your code**. This is handy for making notes to yourself and you can even add hashtags after lines of runable code, on the same line.  

The name of your script file is in the tab at the top of your script window - the name defaults to `Untitled1`. Be sure to save your script by clicking "File --> Save" or `command + s` (Mac) or `Ctrl + s` (PC). You can also click the floppy disk icon to save. 

## Bottom left pane (output/console)

Code output is displayed in the `console` in the bottom left pane. This space is also good for just noodling around and trying out code that you do not wish to save in your script.  

In the console, the prompt `>` looks like a greater than symbol. If your prompt begins to look like a `+` symbol by mistake, simply click in your console and press the `esc` key on your keyboard as many times as necessary to return to the prompt. 

R uses `+` when code is broken up across multiple lines and R is still expecting more code. A line of code does not usually end until R finds an appropriate stop parameter or punctuation that completes some code such as a closed round parenthesis `)`, square bracket `]`, curly brace `}`, or quotation mark `'`.  

If the output in your console gets too messy, you can clear it by pressing `control + l` on both Mac and PC. This will not erase any saved data - it will simply make your console easier to read. 

## Top right pane (global environment)

Data are saved in R's memory as "variables". Variables are simply placeholders for a value, mathematical expression, word, function, or dataset! The global "Environment" tab in the upper right pane displays the variables you have assigned/saved. "Global" simply means that these variables are available for any task.  

## Bottom right pane (files, plots, packages, and help)

Here you find useful tabs for navigating your file system, displaying plots, installing packages, and viewing help pages. Press the `control` key and a number (1 through 9) on your keyboard to shortcut between these panes and tabs. 

# Package installation

Many people write clever software that makes R smarter/better/faster/stronger. You can install packages by clicking the "Packages" tab in the bottom-right pane, clicking install, and then searching for the package you wish to install (or use the `install.packages()` function  - but wrap the name of the package in quotation marks. It is generally good to keep RStudio and your packages up to date. Install the packages we will use in the workshop: 

```{r}
install.packages(c("ggplot2", "knitr", "psych", "rmarkdown", "reshape2", "swirl"), dependencies = TRUE)
```

## The `library` function  

However, installing a package is only the first step of the two-step process of using add-on packages. Installing the package physically puts the files on your computer, but RStudio does not yet know they are there. Anytime you want to utilize a previously installed package, you must retreive it with `library()` when you begin a new R session. You do not need to reinstall packages each time you quit and restart your R session. 

```{r}
library(ggplot2)
library(knitr)
library(psych)
library(rmarkdown)
library(reshape2)
# library(swirl)
# These packages can now be used in our RStudio session!
```

# Getting help  

Typing the `?` symbol before a built-in function or dataset name can be used to bring up the help pages for that particular thing. Below are a few that we will use in this workshop series! 

```{r}
?read.csv
?class
?hist
?plot
?ggplot2::ggplot
?t.test
?lm
?summary
?psych::describe
?iris
```

# Functions and arguments

Our coding universe consists of **functions** and **arguments**.  

* Functions perform actions on things. In the natural language analogy, functions are the verbs in our sentences. An easy way to identify a function is because it contains trailing parentheses such as after `mean()`. 

* Arguments are the things; they consist of virtually anything like values, expressions, vectors, lists, matrices, words, text of entire books, functions, and even large datasets. These arguments go inside the parentheses of functions!

```{r}
# What is the arithmetic mean of 100, 1000, and 10000?
mean(c(100, 1000, 10000))
```

# **Challenge 1**
What are functions and arguments? Describe a few examples of each.  

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>