---
title: 'R Fundamentals Part1-6: lists and matrices'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# List

A **LIST** is an ordered group of data that are not of the same type. Lists are heterogenous. Instead of using `c()` like in vector composition, use `list()` to create a list:

```{r, eval=FALSE}
?list
```
```{r}
list1 = list(TRUE, "one", 1) # include three kinds of data: logical, character, and integer
list1
class(list1)
```

Lists are simple containers and are not additive or multiplicative like vectors and matrices are:
```{r, eval=FALSE}
list1 * list(FALSE, "zero", 0) # Error
```

# Matrix  

**MATRICES** are homogenous like vectors. They are tables comprised of data all of the same type. Matrices are organized into rows and columns. 

Use `matrix()` to create a matrix

```{r, eval=FALSE}
?matrix
```

We can also specify how we want the matrix to be organized using the `nrow` and `ncol` arguments:

```{r}
# Construct a 4 x 3 matrix
matrix1 = matrix(data = 1:12, nrow = 4, ncol = 3) 
matrix1
class(matrix1)
```

We can also coerce a vector to a matrix, because a vector is comprised of homogenous data of the same kind, just like a matrix is:

```{r}
# Create a numeric vector from 1 to 20
vec1 = c(1:20)
vec1
class(vec1)

# Coerce this vector to a matrix with 10 rows and 2 columns:
matrix2 = matrix(vec1, ncol = 2)
matrix2
class(matrix2)
```

> 

> It is worth knowing some basics of lists because in Part3 you will see that statistical output is often returned as a list. Matrices are useful for doing complex math and statistical related-stuff. However, beginners should keep their focus on vectors and dataframes instead of lists and matrices.  

# **Challenge 9**
Read ahead to the Part2 materials to learn about subsetting lists and matrices. In 2-3 minutes of reading/skimming, how do you think subsetting works? 

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>