---
title: 'R Fundamentals Part1-5: data frames'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: html_document
---

# Data frame  

_**It is worth emphasizing the importance of data frames in R!**_ Inside R, a **DATA FRAME** is a list of equal-length vectors. These vectors can be of different types. Data frames are thus heterogenous like lists. Data frames have the **same type of data within each column**, but **different types of data across columns!** This is simply a tabular spreadsheet of rows and columns.  

Let's create a dataframe called `gdp` using the vectors we already created:

```{r}
set.seed(1)
(uniform = runif(n = 10, min = 100, max = 1000000))
(normal = rnorm(n = 10, mean = 0, sd = 1))
(integer = sample(x = 1990:2010, size = 10, replace = FALSE))
(character = sample(c("Canada", "USA", "Mexico"), 10, replace = TRUE))
(logical = sample(c(TRUE, FALSE), 10, replace = TRUE))
```

We do this using `data.frame()`

```{r eval = F}
?data.frame
```
```{r}
gdp = data.frame(uniform, normal, integer, character, logical, 
                 stringsAsFactors = FALSE)
# NOTE: `stringsAsFactors = FALSE` means that R will NOT automatically coerce character vectors to factor/categorical type. 
```

Take a peek at the `gdp` data frame to see what it looks like:

```{r}
gdp
```

We can change the names of the columns by defining them as a vector with our desired names, of equal length to the existing column names (five): 

```{r}
# Create a vector called `new_df_names` with the new column names:
new_col_names = c("gdpPercap", "growth_rate", "year", "country", "healthy")

# Pass this vector into `colnames()`
colnames(gdp) = new_col_names
gdp
```

We can check the structure of our data frame via `str()`

```{r, eval=FALSE}
?str
```
```{r}
str(gdp)
```

Here, we can see that a data frame is just a list of equal-length vectors! For readability purposes, `str()` displays columns in what look like rows, from top to bottom, while the data are displayed left to right. 

## Learning about your data frame  

```{r}
# View the dimensions (nrow x ncol) of the data frame:
dim(gdp) 

# View column names:
colnames(gdp)

# View row names (we did not change these and they default to character type)
rownames(gdp)
class(rownames(gdp))
```

# Change column order

You can change the order of the columns by specifying their new order using `c()` within what is called "bracket notation" `[ , ]`. 

> This will be covered with the rest of subsetting in Part 2. 

```{r}
gdp = gdp[,c("country", "year", "gdpPercap", "growth_rate", "healthy")]
gdp
```

## Sorting data

You can also sort data by increasing/decreasing alphanumeric order. Sort by Country and then Year by typing: 

```{r}
?order
```
```{r}
gdp_sorted = gdp[order(gdp$country, gdp$year),]
gdp_sorted
```

## Factor data type  

Factor data are categorical types used to make comparisons between other data. Factors group the data by their "levels" (the different categorical groups within a factor).  

For example, in our `gdp` dataframe, we can coerce "country" from character to factor type. Canada, Mexico, and USA are the factor levels, or groups. If we want to compare `population` and `growth_rate` between these three countries, we should set the character "country" vector to factor data type. We can do so with `factor()`.  

The dollar sign operator `$` is used to call a single vector from a data frame. This will be discussed more in Part 2 along with bracket notation.  

```{r}
str(gdp) # "country" is character type. 
gdp$Country = factor(gdp$country) # Convert "Country" to factor type. 
str(gdp) # "Country" is now factor type!
```

Notice that R stores factors internally as integers and uses the character strings as labels. Also notice that by default R orders factors alphabetically and returns them when we call the "country" vector.

```{r}
gdp$Country
levels(gdp$Country)
```

### Changing factor "levels"

If we want to change how R stores the factor levels, we can specify them by using the `levels()` argument. For example:

```{r}
gdp$country  # Levels: Canada, Mexico, USA (default alphabetical sort)

# What if we want to change the factor level sort to Levels: Mexico, USA, Canada?

gdp$country = factor(gdp$Country, levels = c("Mexico", "USA", "Canada"))
gdp$country # The level order has changed. 
```

> NOTE: we will load an actual subset of the `gdp` data frame from file at the beginning of Part 2, so do not worry if your dataframe is not identical to your neghbor's! 

# **Challenge 8**
Where have you seen something that looks like a dataframe before?  

[Read this page](https://www.stat.berkeley.edu/~s133/factors.html) for more help with factors.  

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>