---
title: 'R Fundamentals Part4-5: the birthday problem'
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: 
  html_document: 
    number_sections: yes
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---

# The Birthday Problem

# **Challange 6**
Say we choose 25 people at random. What is the probability two or more of them have the same birthday?'

Model simplifications:  

* Ignore leap years.  

* Assume the probability of having a birthday in any of the 365 days of the year is equally likely (and no twins, triplets, etc. in the room).  

Let's start by thinking of one room. We need to sample 25 birthdays and check how many are repeated: 

```{r}
# This seed give one matching birthday
set.seed(94705)  

birthdays = sample(1:365, 25, replace = TRUE)
birthdays
```

We now want to know how many of those birthdays are repeated. For that we can use the `unique()` command.

```{r}
unique(birthdays)
length(unique(birthdays))
```

24 birthdays are unique. Two people have the same birthday!  

Note this is just one realization of the process. To use R to approximate the probability we would need to repeat this process many, many times. 

For this we can start by writing a function that goes through the process one time and then use the `replicate` function as we did above. 

We start by writing the function:

```{r}
set.seed(94705)

birthday_function = function(people = 25){
    
    # we populate the room
    birthdays = sample(1:365, people, replace = TRUE)
    
    # get the unique number of bdays
    unique_bdays = length(unique(birthdays))
    
    # and return a 1 if at least one bday is repeated.
    as.numeric(unique_bdays != people)
}

birthday_function(people = 25)
```

Thus, we have one repeat birthday! 

Now we can use `replicate` to repeat the process 1,000 times:

```{r}
set.seed(94705)
many_sims = replicate(1000, birthday_function(people = 25)) 
```
```{r, eval=FALSE}
many_sims
```

To approximate the probability of at least one matching birthday, we can just take the mean of this vector:

```{r}
mean(many_sims) # 0.578 chance of one matching birthday! 
```

Say we now want to use R to see how this probability changes as the number of people in the room changes. How can we do this? We can use our function inside of a loop to vary the number of people in the room:

```{r}
people = 2:100

sims = matrix(NA, nrow = length(people), ncol = 2)

for (i in 1:length(people)) {
    many_sims = replicate(5000, birthday_function(people = people[i]))
    sims[i,] = c(people[i], mean(many_sims))
}

sims
```

Plot the probability: 

```{r}
# Change our plotting area to a 1 x 1 surface
par(mfrow = c(1,1))

plot(x = sims[,1], y = sims[,2], 
     pch = 16, col = "blue", 
     xlab = "Nr of People", 
     ylab = "Probability of at Least One Match")
```

Learn more about the birthday problem [here](http://mathforum.org/dr.math/faq/faq.birthdayprob.html)

Although the "purrr" R package is perhaps more useful for quickly applying functions to a vector or a list, there exists a whole family of functions for this sort of thing in base R. Learn more about the "apply" functional family [here](https://www.r-bloggers.com/using-the-apply-family-of-functions-in-r/), [here](https://www.r-bloggers.com/using-apply-sapply-lapply-in-r/), and [here](http://adv-r.had.co.nz/Functionals.html).  

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>