---
title: "R Fundamentals Part4-2: for loops"
author: "D-Lab"
date: "March 31, 2019 (updated `r Sys.Date()`)"
output: 
  html_document: 
    number_sections: yes
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Recall that in R, commands we type such as `nrow()` are functions. Learning to write code inside of your own custom functions can potentially save you time and stress.  

What do you think `nrow` does? Type a function name without the parentheses to see its source code.  

```{r, eval=F}
nrow # returns the number of rows in a data frame: dim(x)[1L]
```

Note that doing this manually or via nrow gives the same result:

```{r}
data(iris)

dim(iris)[1L]

nrow(iris)

# Function method is equivalent to manual method
dim(iris)[1L] == nrow(iris) 
```

We can even write our own *custom* function to do the same thing using the same code: 

```{r}
nrow_function = function(x){
  dim(x)[1L]
}
nrow_function(iris)
nrow_function(mtcars)
```

Before we get to functions however, check out for loops first and understand why they are useful.  

# For loops

A for loop repeats a block of code a number of times until a certain condition is met (or is no longer met) which tells the code to stop and complete its process and perhaps return some sort of output.  

The basic syntax looks like this: `for(variable in sequence) {statement}`

1. For loops begin with the `for` function to tell R that you want to iterate over a loop.  

2. The `variable` in `(variable in sequence)` is generally denoted with an `i`, which stands for "iterator". However, `i` should be thought of as a placeholder and can be anything (e.g., `x`, `donut`, etc.). 

3. `sequence` is a number telling R how many times you want to iterate the code. 

4. `{statement}` refers to the code sequence that you want to run at each iteration `i`. Curly braces { } define the statement boundary as the body of the loop.  

## For-loops - Example 1

Let's start simple! If you want to use a for loop to print the numbers one through ten, it might loop like this.  

`x` is the iterator/placeholder variable.  
`in` indicates we want to do the same thing to `x` one through ten times.  
`print(x)` prints the loop's instructions defined by `x` and `in` and returns the printed values.  

```{r}
for (x in 1:10) {
  print(x)
}
```

## For-loops - Example 2

Now, let's perform some sort of calculation in our loop. Create a matrix of 2 to the power of `i` where `i` is 1 to 10. First we create an NA vector called `vec` which will serve as a placeholder for the output of the loop.

```{r}
vec = c(rep(NA, 10))
vec
```

Now, let's use a for-loop to create a matrix of 2 to the power of i where i is 1 to 10:

```{r}
for(i in 1:10){
    vec[i] = 2^(i)
}  # does indentation matter? 
vec
class(vec)
```
This code tells R that we want to raise 2 to the power of `i`, where `i` is 1:10 consecutive iterations.

## Indexing for-loops

We can also use indexing to modify only some elements of a loop. We now have an object called `vec` with defined values. What if we want to replace the first 5 elements of `vec` with 3 to the power of `i` instead?  

You can change the `sequence` to tell R just to overwrite the first five positions!  

```{r}
for (i in 1:5) {
    vec[i] = 3^i
}                
vec
#This has changed only the first five entries of `vec`
```

###For-loops - Example 3

You can also loop over other data types. Create a name vector:

```{r}
animal_names = c("Cat", "Dog", "Pig", "Elephant", "Giraffe")
animal_names
```

Then, create an NA vector of the same length like we did above:

```{r}
animals_length = rep(NA, length(animal_names))
animals_length
```

Now give the NA vector names:

```{r}
names(animals_length) = animal_names
animals_length  
```

Finally, perhaps you want to perform some operation across `animals_length`. For example, count the number of characters in each animal name like this:

```{r}
for (i in animal_names) {
  animals_length[i] = nchar(i)
}
animals_length
```

## For loops (`if` and `else`)  

For long loops, many intermediate commands called "control structures" can help us apply conditions to different parts of the data. Two basic ones are `if` and `else`.  

Refer to the first for loop you wrote above: 

```{r, eval=FALSE}
for (x in 1:10) {
  print(x) 
}
```

### `if`

`if` statements are frequently used to specify code to be evaluated when some condition is held. Here we tell the code that when `x` is no longer less than 5, stop printing  
```{r}
for (x in 1:10) {
  if (x < 5)
    print(x)
}
```

# **Challenge 2**

1. Write a for loop that outputs something. 

```{r}
## YOUR CODE HERE
```

### `else`

Recall that `if` maintains some condition until it is no longer satisfied:  

`if` something happens, do "this",  

... however, `else` can be used to specify what the code should do when that `if` code is no longer satisfied:  

`if` something `else` happens, do "that"!  

syntax: `if (condition) {statement} else {other statement}`

Let's write a loop where `x` will be printed as long as it is greater than seven. If `x` is not greater than seven, the else part will be printed "Error: number not big enough to print".  

Define `x` as 1 and try it: 

```{r}
x = 1
if (x > 7) {
  print(x)
}else{   #`else` should not start its own line. Always let it be preceded by a closing brace on the same line.
  print("Error: number not big enough to print")
}
```

Reassign `x` as 8 here - its value is printed because it satisfies the `if` condition:

```{r}
x = 8
if (x > 7) {
  print(x)
}else{
  print("Error: number not big enough to print")
}
```

This also works in a loop. Here, we get all outputs for the loop when `x` is again defined as the vector 1:10:

```{r}
x = 1:10
for (x in 1:length(x)) {
    if (x > 7) {
        print(x)
    }else{                     
      print("Error: number not big enough to print. 'x' must be greater than 7")
    }
}
#Super cool! :)  This is essentially what R is doing under the hood. 
```

### `ifelse`

The `ifelse` function can be handy to recode data into mutually exclusive categories:  

syntax: `ifelse(test, yes, no)`  

Think back to the random sampling portion of part 1. Generate ten random draws of two countries "Canada" and "Mexico":
```{r}
set.seed(1)
countries = sample(c("Canada", "Mexico"), 10, replace = TRUE)
countries
```

Now we can recode these types of animals to other types of data: integer, factor, etc. Let's use `ifelse` to change "Canada" and "Mexico" to "Snow?" and "Sun!" instead ...

```{r}
weather = ifelse(countries == "Canada", "Snow?", "Sun!")       
weather
```

This works for other data types as well. 

# **Challenge 3**

1. Write a for loop that uses `if` and `else`

```{r, include = F}
## YOUR CODE HERE
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>